---
title: "Phylo spatial meta-analysis"
date: "`r Sys.Date()`"
# author: " "
format: 
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-float: true
    toc-title: "**CONTENTS**"
    output-file: "AM_worked_example.html"
    embed-resources: true
    # code-fold: true
    code-tools: 
      source: false
    number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "12"
    mainfont: "Roboto"
    monofont: "Roboto"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: united
    code-line-numbers: false
    grid:
      margin-width: 180px
crossref: 
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
  # cache: true
---

# Setup

# Phylogenetic meta-analysis

## How to make phylogenetic correlation matrix?
```{r}

```

### Using a specific phylogenetic tree {#using-a-specific-phylogenetic-tree}
```{r}

```

### Using the `rotl` package (Open Tree of Life) {#using-the-rotl-package-open-tree-of-life}
```{r}

```

## Examples

::: panel-tabset
### 1

#### Brownian motion

##### metafor

##### brms
```{r}

```

##### glmmTMB
```{r}

```

#### Ornstein-Uhlenbeck

##### metafor
```{r}

```

### 2

#### Brownian motion

##### metafor
```{r}

```

##### brms
```{r}

```

##### glmmTMB
```{r}

```

#### Ornstein-Uhlenbeck

##### metafor
```{r}

```

:::

# Spatial meta-analysis

## How to make spatial correlation matrix?

## Examples

::: panel-tabset
### 1. Grau-Andrés et al. (2024)

```{r}

dat_Roger <- read.csv(here("data", "examples", "Roger_etal_2024", "Roger_etal_2024.csv"))
head(dat_Roger)

dat_Roger$const <- 1 # add constant for spatial models
dat_Roger$effect_id <- seq_len(nrow(dat_Roger))

dat_Roger <- dat_Roger %>%
  filter(!is.na(longitude))
coords2 <- cbind(dat_Roger$longitude, dat_Roger$latitude)

# project to a planar coordinate system and convert to kilometers
dat_sf2 <- st_as_sf(dat_Roger, coords = c("longitude", "latitude"), crs = 4326)
dat_sf2_proj <- st_transform(dat_sf2, crs = 3857)

coords_m <- st_coordinates(dat_sf2_proj)
dat_Roger$x_km <- coords_m[,1] / 1000
dat_Roger$y_km <- coords_m[,2] / 1000

coords_km <- cbind(dat_Roger$x_km, dat_Roger$y_km)
dist_matrix_euclid2 <- as.matrix(dist(coords_km))

# row and column names
rownames(dist_matrix_euclid2) <- dat_Roger$effect_id
colnames(dist_matrix_euclid2) <- dat_Roger$effect_id

dat_Roger <- dat_Roger %>%
  group_by(latitude, longitude) %>%
  mutate(site_id = cur_group_id()) %>%
  ungroup()
names(dat_Roger)
```

#### Exponential kernel

##### metafor
```{r}

```
##### brms
```{r}

```
##### glmmTMB
```{r}

```

#### Gaussian process

##### metafor
```{r}

```
##### brms
```{r}

```

### 2. Maire et al. (2019)
We use a dataset from the `metadat` package as our second illustrative example. In the original study, spatial correlations among sampling stations were modelled using station-level coordinates along river networks. 
However, precise latitude and longitude information for individual stations is not available in the published dataset; only the locations of study sites can be geo-referenced. 
We therefore used site-level coordinates as proxies for station-level positions. Although this approximation allows us to compare the implementation of spatial meta-analytic models across `metafor`, `brms`, and `glmmTMB`, the resulting estimates are not directly comparable to those reported by Maire et al. (2019), where station-level spatial random effects were explicitly modelled.

```{r}
dat <- dat.maire2019$dat
dmat <- dat.maire2019$dmat
dmat[1:35,1:35]

sites <- tribble(
  ~site_id, ~site_name,   ~lat,    ~lon,
  "site1",  "Belleville",  47.5,    2.65,
  "site2",  "Bugey",       45.8,    5.27,
  "site3",  "Chinon",      47.2,    0.18,
  "site4",  "Chooz",       50.1,    4.80,
  "site5",  "Civaux",      46.4,    0.65,
  "site6",  "Cruas",       44.6,    4.76,
  "site7",  "Dampierre",   47.7,    2.51,
  "site8",  "Nogent",      48.5,    3.52,
  "site9",  "St Alban",    45.3,    4.75,
  "site10", "St Laurent",  47.7,    2.53,
  "site11", "Tricastin",   44.3,    4.73
)


dat_maire2019 <- dat %>%
  left_join(sites, by = c("site" = "site_id")) |> 
  mutate(id = row_number())

dat_maire2019$const <- 1 # add constant for spatial models

# project to a planar coordinate system and convert to kilometers
dat_sf <- st_as_sf(dat_maire2019, coords = c("lon", "lat"), crs = 4326)
dat_sf_proj <- st_transform(dat_sf, crs = 3857)

coords_m <- st_coordinates(dat_sf_proj)
dat_maire2019$x_km <- coords_m[,1] / 1000
dat_maire2019$y_km <- coords_m[,2] / 1000

coords_km <- cbind(dat_maire2019$x_km, dat_maire2019$y_km)
dist_matrix_euclid <- as.matrix(dist(coords_km))

# row and column names
rownames(dist_matrix_euclid) <- dat_maire2019$id
colnames(dist_matrix_euclid) <- dat_maire2019$id

head(dat_maire2019)
```

#### Exponential kernel
##### metafor
```{r}
#| eval: false

ma_sp2_exp <- rma.mv(s1, vars1, 
             random = list(
                       ~ 1|id,
                       ~ 1|site,
                       ~ x_km + y_km | const), 
             struct = "SPEXP", 
             data = dat_maire2019,
             test = "t",
             method = "REML",
             sparse = TRUE,
             verbose = TRUE
             )

summary(ma_sp2_exp)
# Multivariate Meta-Analysis Model (k = 35; method: REML)
# 
# logLik   Deviance        AIC        BIC       AICc   
# -199.0469   398.0939   408.0939   415.7257   410.2367   
# 
# Variance Components:
#   
#   estim     sqrt  nlvls  fixed  factor 
# sigma^2.1   819.5017  28.6269     35     no      id 
# sigma^2.2  1653.5229  40.6635     11     no    site 
# 
# outer factor: const        (nlvls = 1)
# inner term:   ~x_km + y_km (nlvls = 11)
# 
# estim     sqrt  fixed 
# tau^2      8548.0233  92.4555     no 
# rho          97.3090              no 
# 
# Test for Heterogeneity:
#   Q(df = 34) = 191.4837, p-val < .0001
# 
# Model Results:
#   
#   estimate       se    zval    pval     ci.lb     ci.ub      
# 190.8255  40.9187  4.6635  <.0001  110.6263  271.0246  ***

confint(ma_sp2_exp)
# estimate  ci.lb     ci.ub 
# sigma^2.1 819.5017 0.0000 4240.3900 
# sigma.1    28.6269 0.0000   65.1183 
# 
# estimate  ci.lb      ci.ub 
# sigma^2.2 1653.5229 0.0000 21698.5229 
# sigma.2     40.6635 0.0000   147.3042 
# 
# estimate  ci.lb        ci.ub 
# tau^2 8548.0233 0.0000 >854802.3263 
# tau     92.4555 0.0000    >924.5552 
# 
# estimate   ci.lb     ci.ub 
# rho  97.3090 <9.7309 >973.0899 

```

```{r}

```

##### brms
```{r}

```

##### glmmTMB
```{r}

dat_maire2019$id <- factor(dat_maire2019$id)
VCV <- diag(dat_maire2019$vars1, nrow = nrow(dat_maire2019)) 
rownames(VCV)<- colnames(VCV)<- dat_maire2019$id
VCV[1:5, 1:5]

dat_maire2019$pos <- numFactor(dat_maire2019$x_km, dat_maire2019$y_km)

tmb_sp2_exp <- glmmTMB(
  s1 ~ 1 +
    equalto(0 + id | const, VCV) + 
    (1 | site) +
    exp(pos + 0 | const),     
  data = dat_maire2019,
  REML = TRUE
  )


head(confint(tmb_sp2_exp), 10)

sigma(tmb_sp2_exp) # for id

tmb_sp2_exp_varcor <- VarCorr(tmb_sp2_exp)$cond
exp(tmb_sp2_exp$fit$par[[4]]) # for spatial range parameter rho


```

#### Gaussian process

##### metafor
```{r}

ma_sp2_gau <- rma.mv(s1, vars1, 
                          random = list(
                            ~ 1|id,
                            ~ 1|site,
                            ~ x_km + y_km | const), 
                          struct = "SPGAU", 
                          data = dat_maire2019,
                          sparse = TRUE,
                          verbose = TRUE
)

summary(ma_sp2_gau)
# Multivariate Meta-Analysis Model (k = 35; method: REML)
# 
# logLik   Deviance        AIC        BIC       AICc   
# -199.1047   398.2094   408.2094   415.8412   410.3522   
# 
# Variance Components:
#   
#   estim     sqrt  nlvls  fixed  factor 
# sigma^2.1   805.9835  28.3898     35     no      id 
# sigma^2.2  3557.3247  59.6433     11     no    site 
# 
# outer factor: const        (nlvls = 1)
# inner term:   ~x_km + y_km (nlvls = 11)
# 
# estim     sqrt  fixed 
# tau^2      6637.4779  81.4707     no 
# rho         143.9412              no 
# 
# Test for Heterogeneity:
#   Q(df = 34) = 191.4837, p-val < .0001
# 
# Model Results:
#   
#   estimate       se    zval    pval     ci.lb     ci.ub      
# 186.2795  40.6944  4.5775  <.0001  106.5199  266.0392  *** 

confint(ma_sp2_gau)

```

##### brms
```{r}


```

##### glmmTMB
```{r}

tmb_sp2_gau <- glmmTMB(
  s1 ~ 1 +
    equalto(0 + id | const, VCV) + 
    (1 | site) +
    gau(pos + 0 | const),     
  data = dat_maire2019,
  REML = TRUE
)

head(confint(tmb_sp2_gau), 10)

tmb_sp2_gau_varcor <- VarCorr(tmb_sp2_gau)$cond
exp(tmb_sp2_gau$fit$par[[4]]) # for spatial range parameter rho
```

:::

# Software and package versions

```{r}
#| echo: false

sessionInfo()

```


# References
-   Maire, A., Thierry, E., Viechtbauer, W., & Daufresne, M. (2019). Poleward shift in large-river fish communities detected with a novel meta-analysis framework. Freshwater Biology, 64(6), 1143–1156. <https://doi.org/10.1111/fwb.13291>
-   Grau‐Andrés, R., Moreira, B., & Pausas, J. G. (2024). Global plant responses to intensified fire regimes. Global Ecology and Biogeography, 33(8), e13858. <https://doi.org/10.1111/geb.13858>
-    Viechtbauer, W., White, T., Noble, D., Senior, A., & Hamilton, W. K. (2025). metadat: Meta-analysis datasets (Version 1.5-2) [R package]. <https://github.com/wviechtb/metadat>
