---
title: "Phylo spatial meta-analysis"
date: "`r Sys.Date()`"
# author: " "
format: 
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-float: true
    toc-title: "**CONTENTS**"
    output-file: "AM_worked_example.html"
    embed-resources: true
    # code-fold: true
    code-tools: 
      source: false
    number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "12"
    mainfont: "Roboto"
    monofont: "Roboto"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: united
    code-line-numbers: false
    grid:
      margin-width: 180px
crossref: 
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
  # cache: true
---

In this online tutorial, we demonstrate how to conduct phylogenetic and spatial meta-analyses using R packages such as `metafor`, `brms`, and `glmmTMB`. We provide step-by-step instructions on creating phylogenetic and spatial correlation matrices, fitting models with different correlation structures, and interpreting the results.


# Setup
```{r}

# install.packages("pacman") # if not already installed
pacman::p_load(brms, metafor, metadat, tidyverse, data.table, crayon, here, ape, patchwork,
               dplyr, tidyr, MCMCglmm, bayesplot, tidybayes, posterior, orchaRd,purrr, stringr, readr, lubridate, magrittr, janitor, glmmTMB, rotl, sf)
```

Set up parallel processing for `brms` - please adjust according to your computer's capabilities.

```{r}
#| eval: false

max_cores <- 10
num_chains <- 2
threads_per_chain <- floor(max_cores / num_chains)
options(mc.cores = num_chains) 
```
# Phylogenetic meta-analysis

## How to make phylogenetic correlation matrix?

Before moving the each model, we show how to make phylogenetic correlation matrix. It is important to account for the shared evolutionary history of species. Most common way to do this is by incorporating a phylogenetic correlation matrix into your model. This matrix reflects how closely related species are,and can be derived in two main ways:

1.  [Using a specific phylogenetic tree](#using-a-specific-phylogenetic-tree)\
2.  [Using the `rotl` package (Open Tree of Life)](#using-the-rotl-package-open-tree-of-life)

In the following two sections, we introduce how to make the correlation matrix in two ways. The example below demonstrates how to compute a correlation matrix under the *Brownian motion process*. For the *Ornstein–Uhlenbeck process* and *Phylogenetic Gaussian Process*, please refere to their respective sections.


:::callout-note
Sometimes, the species names in your data may not match those in the tree - please ensure that they are consistent. Also, it will be necessity when dropping species from the tree that are not in your dataset. In that case, you can use `TreeTools::DropTip()` / `ape::drop.tip()` (or `TreeTools::KeepTip()` / `ape::keep.tip()`). You can see how to do this in the example below :)
:::

### Using a specific phylogenetic tree {#using-a-specific-phylogenetic-tree}
In some cases, the phylogenetic relationships among the species of interest are already well established, and a corresponding phylogenetic tree is available from a published source or a public repository. When this is the case, it is usually preferable to use the existing tree rather than constructing a new one automatically. For example, large, time-calibrated phylogenies are available for many taxonomic groups (e.g. [birds](BirdTree.org), mammals, plants), often accompanied by branch lengths and documentation of the underlying assumptions.
Using such trees ensures consistency with previous studies and avoids unnecessary reconstruction steps.

Below, we illustrate how to use a phylogenetic tree provided as a file and prepare it for use in downstream analyses, such as phylogenetic meta-analysis.

#### Step 1. Read the tree from a file
Phylogenetic trees are commonly distributed in formats such as Newick (`.nwk`, `.tre`) or Nexus (`.nex`). The ape package can read most standard formats.
<!-- probably change here - use real data -->
```{r}
#| label: way 1 - phylometa
#| eval: false

library(ape)

# Read a phylogenetic tree from file
# If the file contains multiple trees, select a single representative tree
tree <- read.nexus("tree.txt")

tree
```

#### Step 2. Basic structural checks
Before using the tree in analyses, it is good practice to perform some basic checks to ensure that the tree structure is appropriate.
```{r}
#| eval: false

# Check whether the tree is ultrametric (time-calibrated)
is.ultrametric(tree)

# Check whether the tree is fully bifurcating
is.binary(tree)
```
Phylogenetic comparative and meta-analytic models assume a time-calibrated, fully bifurcating tree. If the tree is not ultrametric, you may need to convert it using methods such as `chronos()` from the `ape` package or `make.ultrametric()` from the `phytools` package. If the tree has polytomies, you can resolve them randomly using `multi2di()` from the `ape` package.

#### Step 3. Match tree tip labels to the dataset
The species names in the tree must match exactly the species identifiers used in the dataset.
```{r}
#| eval: false

# species names used in the dataset
species_data <- unique(dat$species)

# check for mismatches between tree tip labels and dataset species names
setdiff(species_data, tree$tip.label)
setdiff(tree$tip.label, species_data)

```
If mismatches occur, theytypically reflect typographical errors, outdated taxonomy, or diffrences in naming conventions (e.g., underscores vs spaces), These issues must be resolved manually before proceeding. 

#### Step 4. Prune the tree to include only species in the dataset
Phylogenetic trees often contain more species than required for a given analysis, Extraneous tips should be removed to match the dataset.

```{r}
#| eval: false

tree_pruned  <- drop.tip(
  tree,
  setdiff(tree$tip.label, species_data)
  )

tree_pruned

# check the binary and ultrametric status again
is.binary(tree_pruned)
is.ultrametric(tree_pruned)
```

#### Step 5. Compute the phylogenetic correlation matrix
Once the tree is time-calibrated, fully bifurcating, and pruned to match the dataset, you can compute the phylogenetic correlation matrix under a Brownian motion model using the `vcv()` function from the `ape` package.

```{r}
#| eval: false

A <-  vcv(tree_pruned, corr = TRUE)
# check the dimensions and preview the top-left block
dim(A)        
```
This matrix represents the expected correlation among species due to shared evolutionary history and can ve used directly in phylogenetic comparative or meta-analytic models.


### Using the `rotl` package (Open Tree of Life) {#using-the-rotl-package-open-tree-of-life}
Life is not so that easy that you always have a phylogenetic tree for your species of interest. In such cases, you can use the `rotl` package to access the Open Tree of Life database and retrieve a phylogenetic tree based on the *Latin* species names in your dataset.

:::callout-note
- Make sure that the species names in your dataset are in Latin (scientific names) and are spelled correctly. The `rotl` package relies on these names to find the corresponding species in the Open Tree of Life (https://opentreeoflife.github.io/use).
- The OpenTree synthetic tree and name resolution services are actively updated. Running the same code at different times may yield slightly different trees due to updates in the database.
- For reproducible research, it is good practice to save the retrieved tree to a file (e.g., in Newick or Nexus format) after obtaining it using `rotl`. This way, you can ensure that you are using the same tree in future analyses.
- Automated name matching is convenient but may not always be perfect. It is advisable to manually check the matched names to ensure accuracy.
:::

```{r}
#| eval: false

# install and load necessary packages - if not already installed
# install.packages("rotl")
# install.packages("ape")

library(rotl)
library(ape)

# record info about OpenTree version
tol_about()
# OpenTree Synthetic Tree of Life.

# Tree version: opentree16.1
# Taxonomy version: 3.7draft3
# Constructed on: 2025-12-20 00:55:58
# Number of terminal taxa: 2385875
# Number of source trees: 2064
# Number of source studies: 1931
# Source list present: false
# Root taxon: cellular organisms
# Root ott_id: 93302
# Root node_id: ott93302
```

#### Step 1. Provide a species list
Here we use 10 commpnly used lab model taxa across different animal groups as an example. Use Latin names, not common names. 

```{r}
#| eval: false

myspecies <- c(
  "Escherichia colli",                # typo on purpose
  "Chlamydomonas reinhardtii",
  "Drosophila melanogaster",
  "Arabidopsis thaliana",
  "Rattus norvegicus",
  "Mus musculus",
  "Cavia porcellus",
  "Xenopus laevis",
  "Saccharomyces cervisae",           # typo on purpose
  "Danio rerio"
)
```

#### Step 2. Resolve species names using OpenTree TNRS
We first try a strict match, and then a fuzzy match for any unmatched names. This is useful to detect typos or synonyms in species names.

```{r}
#| eval: false

taxa_strict <- tnrs_match_names(
  names = myspecies, 
  do_approximate_matching = FALSE
  )
#   Warning message:
# Escherichia colli, Saccharomyces cervisae are not matched 

taxa_strict
#                search_string               unique_name approximate_match score
# 1          escherichia colli                      <NA>                NA    NA
# 2  chlamydomonas reinhardtii Chlamydomonas reinhardtii             FALSE     1
# 3    drosophila melanogaster   Drosophila melanogaster             FALSE     1
# 4       arabidopsis thaliana      Arabidopsis thaliana             FALSE     1
# 5          rattus norvegicus         Rattus norvegicus             FALSE     1
# 6               mus musculus              Mus musculus             FALSE     1
# 7            cavia porcellus           Cavia porcellus             FALSE     1
# 8             xenopus laevis            Xenopus laevis             FALSE     1
# 9     saccharomyces cervisae                      <NA>                NA    NA
# 10               danio rerio               Danio rerio             FALSE     1
#     ott_id is_synonym flags number_matches
# 1       NA         NA  <NA>             NA
# 2    33153      FALSE                    1
# 3   505714      FALSE                    1
# 4   309263      FALSE                    1
# 5   271555      FALSE                    1
# 6   542509      FALSE                    1
# 7   744000      FALSE                    1
# 8   465096      FALSE                    1
# 9       NA         NA  <NA>             NA
# 10 1005914      FALSE                    1

taxa <- tnrs_match_names(
  names = myspecies,
  do_approximate_matching = TRUE
)

taxa
#                search_string               unique_name approximate_match
# 1          escherichia colli          Escherichia coli              TRUE
# 2  chlamydomonas reinhardtii Chlamydomonas reinhardtii             FALSE
# 3    drosophila melanogaster   Drosophila melanogaster             FALSE
# 4       arabidopsis thaliana      Arabidopsis thaliana             FALSE
# 5          rattus norvegicus         Rattus norvegicus             FALSE
# 6               mus musculus              Mus musculus             FALSE
# 7            cavia porcellus           Cavia porcellus             FALSE
# 8             xenopus laevis            Xenopus laevis             FALSE
# 9     saccharomyces cervisae  Saccharomyces cerevisiae              TRUE
# 10               danio rerio               Danio rerio             FALSE
#        score  ott_id is_synonym          flags number_matches
# 1  0.9375000  474506      FALSE sibling_higher              2
# 2  1.0000000   33153      FALSE                             1
# 3  1.0000000  505714      FALSE                             1
# 4  1.0000000  309263      FALSE                             1
# 5  1.0000000  271555      FALSE                             1
# 6  1.0000000  542509      FALSE                             1
# 7  1.0000000  744000      FALSE                             1
# 8  1.0000000  465096      FALSE                             1
# 9  0.9166667  356221      FALSE                             2
# 10 1.0000000 1005914      FALSE                             1
```
If the strict match fails or returns NAs, you can enable approximate matching. Approximate matching uses fuzzy string matching to find the closest matches in the OpenTree taxonomy (but you should always check the returned matches).

Key columns to check include:
- `unique_name`: the matched OpenTree taxon name
- `ott_id`: the OpenTree taxonomy identifier used for tree retrieval
- `approximate_match`: `TRUE` indicates a non-exact match (for example due to a typo)
- `is_synonym`: `TRUE` indicates the input name was treated as a synonym

A practical rule of thumb is the following:
- If `approximate_match == TRUE` and the reason is an obvious typo (e.g. misspelling of a well-known species name), correct the original species list and re-run the name matching. This ensures that the final dataset contains clean and unambiguous names.
- If a match looks suspicious (for example, the matched taxon belongs to a different genus, or the name could plausibly refer to multiple taxa),
do not accept the match blindly. Instead, inspect the additional information returned by `tnrs_match_names()` to confirm that the matched taxon is indeed the one you intended to include.

#### Step 3. Fix typos and re-run matching

```{r}
#| eval: false

myspecies_fixed <- c(
  "Escherichia coli",
  "Chlamydomonas reinhardtii",
  "Drosophila melanogaster",
  "Arabidopsis thaliana",
  "Rattus norvegicus",
  "Mus musculus",
  "Cavia porcellus",
  "Xenopus laevis",
  "Saccharomyces cerevisiae",
  "Danio rerio"
)

taxa_fixed <- tnrs_match_names(
  names = myspecies_fixed,
  do_approximate_matching = FALSE
)

taxa_fixed
```

At this point, you should confirm that:
-	All species have a valid ott_id.
-	The returned matches correspond to the intended taxa.

#### Step 4. Retrieve the phylogenetic tree from OpenTree
Once you are satisfied with the matched OpenTree identifiers, you can request a trimmed subtree from the OpenTree synthetic tree using `tol_induced_subtree()`.

```{r}
#| eval: false

tree <- tol_induced_subtree(
  ott_ids = taxa_fixed[["ott_id"]],
  label_format = "name"
) 
# we can get a Warning meassage here about collapsing single nodes - this is fine.

# the tree is returned as an ape::phylo object, which can be manipulated and plotted using functions from the ape package.
plot(tree, cex=.8, label.offset =.1, no.margin = TRUE)
```

#### Step 5. Clean and standardise tip labels
OpenTree tip labels can include underscores, and sometimes a tip represents a non-taxon internal node (such as an MRCA node). This often happens when only one representative from a large clade is included (e.g., a single bacterial species).

In our example, you can see that species names have OpenTree identifiers appended, and E. coli is replaced with “mrcaott616ott617”, a non-taxon node within Bacteria. Since there are no other bacteria in the data set and the tree position is correct, we can simply rename the tip back to “Escherichia coli.”

```{r}
#| eval: false

tree$tip.label # see the current tree tip labels
#  [1] "Arabidopsis_thaliana"      "Chlamydomonas_reinhardtii"
#  [3] "Mus_musculus"              "Rattus_norvegicus"        
#  [5] "Cavia_porcellus"           "Xenopus_laevis"           
#  [7] "Danio_rerio"               "Drosophila_melanogaster"  
#  [9] "Saccharomyces_cerevisiae"  "mrcaott616ott617"   

# Replace underscores with spaces for readability
tree$tip.label <- gsub("_", " ", tree$tip.label)

# Relabel MRCA-like tips for didactic plotting purposes
ecoli <- tol_node_info(ott_id = 474506, include_lineage = TRUE)
tax_lineage(ecoli) # domain within Bacteria
#      rank               name        unique_name ott_id
# 1  domain           Bacteria           Bacteria 844192
# 2 no rank cellular organisms cellular organisms  93302

tnrs_match_names(names = ("Escherichia"))  # we can see what info is available for this genus - not much, looks like a messy bit of a tree flagged as "INCONSISTENT"
tree$tip.label <- gsub("mrcaott616ott617", "Escherichia coli", tree$tip.label) # replace the tip label back with our species name
```

It is time to plot and tidy up the tree (note no branch lengths are included, they can be created later via simulations).
```{r}
#| eval: false

plot(tree, cex=.8, label.offset =.1, no.margin = TRUE) 
```

#### Step 6. Final checks (binary tree and matching tip labels)
Before using the tree in downstream analyses, it is good practice to confirm that:
1. The tree is fully binary (no polytomies), if that is required for your method.
2. The tip labels match exactly the species list in your dataset.
```{r}
#| eval: false

# check whether the tree is binary
is.binary(tree) # should return TRUE
# [1] TRUE

# check exact matching between the species list and tree tip labels
intersect(as.character(tree$tip.label), myspecies_fixed)
#  [1] "Arabidopsis thaliana"      "Chlamydomonas reinhardtii"
#  [3] "Mus musculus"              "Rattus norvegicus"        
#  [5] "Cavia porcellus"           "Xenopus laevis"           
#  [7] "Danio rerio"               "Drosophila melanogaster"  
#  [9] "Saccharomyces cerevisiae"  "Escherichia coli"  

setdiff(myspecies_fixed, as.character(tree$tip.label))
# character(0)
setdiff(as.character(tree$tip.label), myspecies_fixed)
# character(0)

```

#### Step 7. Compute the phylogenetic correlation matrix (handling missing branch lengths)

To compute a phylogenetic variance covariance (VCV) or correlation matrix under a Brownian motion (BM) process, we use `vcv()` from the `ape` package. Importantly, `vcv()` requires branch lengths. However, the Open Tree of Life synthetic tree often provides a well-resolved tpology but does not always include branch lengths.

If the tree has no branch lengths, `vcv()` will fail with:
```{r}
#| eval: false

A <- vcv(tree, corr = TRUE)
Error in vcv.phylo(tree, corr = TRUE) : the tree has no branch lengths
```

For didactic purposes in this tutorial, we assign equal branch lengths to all edges (that is, all branch lengths are set to 1). This allows us to demonstrate the workflow and obtain a valid correlation matrix. In real analyses, you should consider using a tree with biologically meaningful branch lengths (for example, a time-calibrated phylogeny).

```{r}
#| eval: false

# check whether the OpenTree synthetic tree includes branch lengths
is.null(tree$edge.length)
# [1] TRUE # TRUE indicates that branch lengths are missing

# if branch lengths are missing, add equal branch lengths for didactic purposes
if (is.null(tree$edge.length)) {
  tree$edge.length <- rep(1, nrow(tree$edge))
  }

# compute the phylogenetic correlation matrix under Brownian motion
A <- vcv(tree, corr = TRUE)

# check the dimensions and preview the top-left block
dim(A)        # should return the number of species (rows/columns)
# [1] 10 10

A[1:5, 1:5]
#                           Arabidopsis_thaliana Chlamydomonas_reinhardtii
# Arabidopsis_thaliana                 1.0000000                 0.6666667
# Chlamydomonas_reinhardtii            0.6666667                 1.0000000
# Mus_musculus                         0.2041241                 0.2041241
# Rattus_norvegicus                    0.2041241                 0.2041241
# Cavia_porcellus                      0.2182179                 0.2182179
#                           Mus_musculus Rattus_norvegicus Cavia_porcellus
# Arabidopsis_thaliana         0.2041241         0.2041241       0.2182179
# Chlamydomonas_reinhardtii    0.2041241         0.2041241       0.2182179
# Mus_musculus                 1.0000000         0.8750000       0.8017837
# Rattus_norvegicus            0.8750000         1.0000000       0.8017837
# Cavia_porcellus              0.8017837         0.8017837       1.0000000
```

## Example

::: panel-tabset
###  Moura et al. (2021)
```{r}
#| eval: false

dat_moura2021 <- dat.moura2021$dat
tree <- dat.moura2021$tree

# calculate r-to-z transformed correlations and corresponding sampling variances
dat_moura2021 <- escalc(measure="ZCOR", ri=ri, ni=ni, data=dat_moura2021)

# turn tree into an ultrametric one
tree <- compute.brlen(tree)

# compute phylogenetic correlation matrix
A <- vcv(tree, corr = TRUE)

# make copy of the species.id variable
dat_moura2021$species.id.phy <- dat_moura2021$species.id

```

#### Brownian motion

##### metafor
```{r}
#| echo: false

# phylo_eg1_meta_BM <- readRDS(here("Rdata", "tutorial_v2", "moura2021_BM_metafor.rds"))

```

```{r}
#| eval: false

system.time(
  phylo_eg1_meta_BM <- rma.mv(yi, vi,
         random = list(~ 1 | study.id, 
         ~ 1 | effect.size.id, 
         ~ 1 | species.id, 
         ~ 1 | species.id.phy),
         R = list(species.id.phy=A), 
         data = dat_moura2021,
         verbose = TRUE,
         sparse = TRUE,
         method = "REML"
         )
  )

summary(phylo_eg1_meta_BM)
# Multivariate Meta-Analysis Model (k = 1828; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -167.6726   335.3453   345.3453   372.8974   345.3782   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed          factor    R 
# sigma^2.1  0.0192  0.1384    457     no        study.id   no 
# sigma^2.2  0.0145  0.1202   1828     no  effect.size.id   no 
# sigma^2.3  0.0557  0.2359    341     no      species.id   no 
# sigma^2.4  0.0512  0.2263    341     no  species.id.phy  yes 
# 
# Test for Heterogeneity:
# Q(df = 1827) = 10743.8076, p-val < .0001
# 
# Model Results:
# 
# estimate      se    zval    pval   ci.lb   ci.ub     
#   0.3682  0.1300  2.8311  0.0046  0.1133  0.6230  ** 

confint(phylo_eg1_meta_BM)
#           estimate  ci.lb  ci.ub 
# sigma^2.1   0.0192 0.0108 0.0325 
# sigma.1     0.1384 0.1038 0.1802 

#           estimate  ci.lb  ci.ub 
# sigma^2.2   0.0145 0.0121 0.0172 
# sigma.2     0.1202 0.1099 0.1311 

#           estimate  ci.lb  ci.ub 
# sigma^2.3   0.0557 0.0334 0.0788 
# sigma.3     0.2359 0.1827 0.2807 

#           estimate  ci.lb  ci.ub 
# sigma^2.4   0.0512 0.0179 0.1792 
# sigma.4     0.2263 0.1336 0.4233 

orchaRd::i2_ml(phylo_eg1_meta_BM)
# I2_Total       I2_study.id I2_effect.size.id     I2_species.id I2_species.id.phy 
# 97.30527          13.26902          10.00808          38.55098          35.47718 

phylo_heritability <- phylo_eg1_meta_BM$sigma2[4] / (phylo_eg1_meta_BM$sigma2[4] + phylo_eg1_meta_BM$sigma2[3])
phylo_heritability
# [1] 0.479239
```

##### brms
```{r}
#| echo: false

# phylo_eg1_brms <- readRDS(here("Rdata", "tutorial_v2", "moura2021_BM_brms2.rds"))

```

```{r}
#| eval: false

fit_phylo_eg1_brms <- bf(yi | se(sqrt(vi)) ~ 1 +
                     (1 | study.id) + 
                     (1 | effect.size.id) + 
                     (1 | species.id) + 
                     (1 | gr(species.id.phy, cov = A))
                     )

prior_brms2 <- get_prior(
  formula = fit_phylo_eg1_brms,
  data = dat_moura2021,
  data2 = list(A = A),
  family = gaussian()
  )

system.time(
  phylo_eg1_brms <- brm(
    formula = fit_phylo_eg1_brms,
    family = gaussian(),
    data = dat_moura2021,
    data2 = list(A = A),
    prior = prior_brms2,
    iter = 10000, 
    warmup = 7000,  
    chains = num_chains,
    backend = "cmdstanr",
    threads = threading(threads_per_chain), 
    control = list(adapt_delta = 0.95, max_treedepth = 15)
    )
  )

summary(phylo_eg1_brms)
#  Family: gaussian 
#   Links: mu = identity 
# Formula: yi | se(sqrt(vi)) ~ 1 + (1 | study.id) + (1 | effect.size.id) + (1 | species.id) + (1 | gr(species.id.phy, cov = A)) 
#    Data: dat_moura2021 (Number of observations: 1828) 
#   Draws: 2 chains, each with iter = 10000; warmup = 7000; thin = 1;
#          total post-warmup draws = 6000
# 
# Multilevel Hyperparameters:
# ~effect.size.id (Number of levels: 1828) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.12      0.01     0.11     0.13 1.00     2091     3894
# 
# ~species.id (Number of levels: 341) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.23      0.03     0.17     0.28 1.00      795     1292
# 
# ~species.id.phy (Number of levels: 341) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.27      0.09     0.14     0.50 1.00     1158      954
# 
# ~study.id (Number of levels: 457) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.14      0.02     0.11     0.18 1.00      661     1223
# 
# Regression Coefficients:
#           Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept     0.37      0.16     0.05     0.70 1.00     3504     2965
# 
# Further Distributional Parameters:
#       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sigma     0.00      0.00     0.00     0.00   NA       NA       NA

```
Let's calculate $I^2$ statistics to quantify the proportion of variance attributable to different sources of heterogeneity.
```{r}
#| eval: false

fit <- phylo_eg1_brms  
dat <- dat_moura2021
v_harm <- 1 / mean(1/dat$vi, na.rm = TRUE)

I2_total_harm  <- var_total_hetero / (var_total_hetero + v_harm)
I2_effect_harm <- var_effect      / (var_total_hetero + v_harm)
I2_study_harm  <- var_study       / (var_total_hetero + v_harm)
I2_species_harm<- var_species     / (var_total_hetero + v_harm)
I2_phylo_harm  <- var_phylo       / (var_total_hetero + v_harm)

round(rbind(
  total   = summ_I2(I2_total_harm),
  effect  = summ_I2(I2_effect_harm),
  study   = summ_I2(I2_study_harm),
  species = summ_I2(I2_species_harm),
  phylo   = summ_I2(I2_phylo_harm)
), 3)
#         mean q2.5.2.5% q50.50% q97.5.97.5%
# total   0.978     0.969   0.978       0.990
# effect  0.090     0.043   0.091       0.134
# study   0.129     0.048   0.123       0.247
# species 0.330     0.115   0.333       0.530
# phylo   0.429     0.178   0.414       0.759

```

##### glmmTMB
```{r}
#| echo: false

# phylo_eg1_tmb <- readRDS(here("Rdata", "tutorial_v2", "phylo_eg1_tmb1.rds"))
```

```{r}
#| eval: false

A <- A[sort(rownames(A)), sort(rownames(A))]

dat_moura2021$g <- 1 

VCV <- diag(dat_moura2021$vi, nrow = nrow(dat_moura2021)) 
rownames(VCV)<- colnames(VCV)<- dat_moura2021$effect.size.id

dat_moura2021$effect.size.id <- as.factor(dat_moura2021$effect.size.id)
head(dat_moura2021)

system.time(
    phylo_eg1_tmb <- glmmTMB(yi ~ 1 + equalto(0 + effect.size.id|g, VCV) +
                         (1|study.id) + 
                         (1|species.id.phy) +
                         propto(0 + species.id.phy|g, A),
                       data = dat_moura2021,
                       REML = T)
    )
head(confint(phylo_eg1_tmb), 10)
#                                                                           2.5 %    97.5 %   Estimate
# (Intercept)                                                            0.1132610 0.6230707 0.3681658
# Std.Dev.(Intercept)|study.id                                           0.1056462 0.1813458 0.1384142
# Std.Dev.(Intercept)|species.id.phy                                     0.1932854 0.2879763 0.2359271
# Std.Dev.species.id.phyAcanthurus_leucosternon_ott388125|g.1            0.1293612 0.3959730 0.2263262
# Std.Dev.species.id.phyAcanthurus_nigricans_ott467313|g.1               0.1293612 0.3959730 0.2263262
# Std.Dev.species.id.phyAcanthurus_nigrofuscus_ott605289|g.1             0.1293612 0.3959730 0.2263262
# Std.Dev.species.id.phyAchatina_fulica_ott997087|g.1                    0.1293612 0.3959730 0.2263262
# Std.Dev.species.id.phyAegithalos_glaucogularis_vinaceus_ott5560982|g.1 0.1293612 0.3959730 0.2263262
# Std.Dev.species.id.phyAethia_pusilla_ott855484|g.1                     0.1293612 0.3959730 0.2263262
# Std.Dev.species.id.phyAgalychnis_callidryas_ott9483|g.1                0.1293612 0.3959730 0.2263262

sigma(phylo_eg1_tmb)^2
# [1] 0.01445014 # residual variance

phylo_eg1_tmb_varcor <- VarCorr(phylo_eg1_tmb)$cond
exp(phylo_eg1_tmb$fit$par)
#   betadisp      theta      theta      theta 
# 0.12020875 0.13841423 0.23592711 0.05122353 <- residual variance (sd), study.id (sd), species.id (sd), species.id.phy (variance?)

```

#### Ornstein-Uhlenbeck

```{r}
#| eval: false

# create phylogenetic distance matrix
dat_moura2021$const <- 1
I  <- matrix(1, nrow = length(tree$tip.label), ncol = length(tree$tip.label))
D <- I-A

D[1:5, 1:5]

system.time(
    phy_ou_meta <- rma.mv(yi, vi,
   random = list(~ 1 | study.id, 
   ~ 1 | effect.size.id, 
   ~ 1 | species.id, 
   ~ species.id.phy | const),
   dist = list(D), 
   struct = "SPEXP",
   control = list(rho.init = 1),
   data = dat_moura2021,
   verbose = TRUE,
   sparse = TRUE,
   method = "REML",
   test = "t"
   )
   )

summary(phy_ou_meta)
# Multivariate Meta-Analysis Model (k = 1828; method: REML)

#    logLik   Deviance       AIC        BIC       AICc   
# -160.3783   320.7567   332.7567   365.8192   332.8028   

# Variance Components:

#             estim    sqrt  nlvls  fixed          factor 
# sigma^2.1  0.0157  0.1254    457     no        study.id 
# sigma^2.2  0.0144  0.1201   1828     no  effect.size.id 
# sigma^2.3  0.0000  0.0000    341     no      species.id 

# outer factor: const           (nlvls = 1)
# inner term:   ~species.id.phy (nlvls = 341)

#             estim    sqrt  fixed 
# tau^2      0.1029  0.3208     no 
# rho        0.0182             no 

# Test for Heterogeneity:
# Q(df = 1827) = 10743.8076, p-val < .0001

# Model Results:

# estimate      se    zval    pval   ci.lb   ci.ub      
#   0.3514  0.0355  9.9090  <.0001  0.2819  0.4209  *** 

# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

rho <- phy_ou_meta$rho  #[1] 0.01820382
alpha <- 1/rho # 54.93351
A_OU <- exp(-alpha*D)

system.time(
  phy_ou_meta1 <- rma.mv(yi, vi,
     random = list(~ 1 | study.id, 
    ~ 1 | effect.size.id, 
    ~ 1 | species.id, 
    ~ 1 | species.id.phy),
    R = list(species.id.phy = A_OU), 
    data = dat_moura2021,
    verbose = TRUE,
    sparse = TRUE,
    method = "REML",
    test = "t"
    )
)

summary(phy_ou_meta1)
# Multivariate Meta-Analysis Model (k = 1828; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -160.3783   320.7567   330.7567   358.3088   330.7896   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed          factor    R 
# sigma^2.1  0.0157  0.1254    457     no        study.id   no 
# sigma^2.2  0.0144  0.1201   1828     no  effect.size.id   no 
# sigma^2.3  0.0000  0.0001    341     no      species.id   no 
# sigma^2.4  0.1029  0.3208    341     no  species.id.phy  yes 
# 
# Test for Heterogeneity:
# Q(df = 1827) = 10743.8076, p-val < .0001
# 
# Model Results:
# 
# estimate      se    tval    df    pval   ci.lb   ci.ub      
#   0.3514  0.0355  9.9090  1827  <.0001  0.2819  0.4210  *** 

```


:::

# Spatial meta-analysis

## How to make spatial correlation matrix?

## Examples

::: panel-tabset
###1. Grau-Andrés et al. (2024)

```{r}
#| eval: false

dat_Roger <- read.csv(here("data", "examples", "Roger_etal_2024", "Roger_etal_2024.csv"))
head(dat_Roger)

dat_Roger$const <- 1 # add constant for spatial models
dat_Roger$effect_id <- seq_len(nrow(dat_Roger))

dat_Roger <- dat_Roger %>%
  filter(!is.na(longitude))
coords2 <- cbind(dat_Roger$longitude, dat_Roger$latitude)

# project to a planar coordinate system and convert to kilometers
dat_sf2 <- st_as_sf(dat_Roger, coords = c("longitude", "latitude"), crs = 4326)
dat_sf2_proj <- st_transform(dat_sf2, crs = 3857)

coords_m <- st_coordinates(dat_sf2_proj)
dat_Roger$x_km <- coords_m[,1] / 1000
dat_Roger$y_km <- coords_m[,2] / 1000

coords_km <- cbind(dat_Roger$x_km, dat_Roger$y_km)
dist_matrix_euclid2 <- as.matrix(dist(coords_km))

# row and column names
rownames(dist_matrix_euclid2) <- dat_Roger$effect_id
colnames(dist_matrix_euclid2) <- dat_Roger$effect_id

dat_Roger <- dat_Roger %>%
  group_by(latitude, longitude) %>%
  mutate(site_id = cur_group_id()) %>%
  ungroup()
names(dat_Roger)
```

#### Exponential kernel

##### metafor
```{r}
#| eval: false

system.time(sp_eg1_exp_mf <- rma.mv(d_Hedges, var_Hedges, 
                              random = list(
                                ~ 1|effect_id,
                                # ~ 1|site_id,
                                ~ x_km + y_km |const
                              ), 
                              struct = "SPEXP", 
                              data = dat_Roger,
                              sparse = TRUE,
                              verbose = TRUE,
                              method = "REML",
                              test = "t"
                              )
            )

summary(sp_eg1_exp_mf)
# Multivariate Meta-Analysis Model (k = 2361; method: REML)
# 
#     logLik    Deviance         AIC         BIC        AICc   
# -4034.4852   8068.9704   8076.9704   8100.0360   8076.9874   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed     factor 
# sigma^2    0.7935  0.8908   2361     no  effect_id 
# 
# outer factor: const        (nlvls = 1)
# inner term:   ~x_km + y_km (nlvls = 383)
# 
#             estim    sqrt  fixed 
# tau^2      1.2330  1.1104     no 
# rho        0.1744             no 
# 
# Test for Heterogeneity:
# Q(df = 2360) = 21273.5149, p-val < .0001
# 
# Model Results:
# 
# estimate      se     tval    df    pval    ci.lb    ci.ub      
#  -0.3349  0.0644  -5.2006  2360  <.0001  -0.4612  -0.2086  *** 

confint(sp_eg1_exp_mf)
#         estimate  ci.lb  ci.ub 
# sigma^2   0.7935 0.7257 0.8678 
# sigma     0.8908 0.8519 0.9316 
# 
#       estimate  ci.lb  ci.ub 
# tau^2   1.2330 1.0147 1.4976 
# tau     1.1104 1.0073 1.2238 
# 
#     estimate   ci.lb  ci.ub 
# rho   0.1744 <0.0174 0.8572 

```

##### brms

```{r}
#| eval: false

dat_Roger$effect_id <- as.character(dat_Roger$effect_id)
vcv <- diag(dat_Roger$var_Hedges)

rownames(vcv) <- colnames(vcv) <- dat_Roger$effect_id

fit_sp_eg1_exp_brms <- bf(d_Hedges ~ 1 +
                 # (1|site_id) + # this is site-level random effect
                 (1|gr(effect_id, cov = vcv)) + # this is m (sampling error)
                gp(x_km, y_km, 
                   cov = "exponential",
                   scale = FALSE)
                )

# generate default priors
prior <- default_prior(fit_sp_eg1_exp_brms, 
                           data = dat_Roger, 
                           data2 = list(vcv = vcv),
                           family = gaussian())
prior$prior[5] = "constant(1)" # meta-analysis assumes sampling variance is known so fixing this to 1
prior
 #                            prior     class       coef     group resp dpar nlpar lb ub tag       source
 #          student_t(3, -0.1, 2.5) Intercept                                                     default
 #                           (flat)    lscale                                       0             default
 # inv_gamma(1.494197, 2141.138175)    lscale gpx_kmy_km                            0             default
 #             student_t(3, 0, 2.5)        sd                                       0             default
 #                      constant(1)        sd            effect_id                  0             default <- check this
 #                      constant(1)        sd  Intercept effect_id                  0        (vectorized) <- check this
 #             student_t(3, 0, 2.5)      sdgp                                       0             default
 #             student_t(3, 0, 2.5)      sdgp gpx_kmy_km                            0        (vectorized)
 #             student_t(3, 0, 2.5)     sigma                                       0             default

# fitting model
sp_eg1_exp_brms <- brm(
  formula = fit_sp_eg1_exp_brms,
  data = dat_Roger,
  data2 = list(vcv=vcv),
  chains = 2,
  iter = 6000,
  warmup = 3000,
  prior = prior,
  control = list(adapt_delta=0.95, max_treedepth=15)
  )
summary(sp_eg1_exp_brms)
#  Family: gaussian 
#   Links: mu = identity 
# Formula: d_Hedges ~ 1 + (1 | gr(effect_id, cov = vcv)) + gp(x_km, y_km, cov = "exponential", scale = FALSE) 
#    Data: dat_Roger (Number of observations: 2361) 
#   Draws: 2 chains, each with iter = 6000; warmup = 3000; thin = 1;
#          total post-warmup draws = 6000
# 
# Gaussian Process Hyperparameters:
#                    Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sdgp(gpx_kmy_km)       1.14      0.07     1.00     1.29 1.00     1818     3552
# lscale(gpx_kmy_km)   128.18     26.02    85.59   187.10 1.00     2227     2567
# 
# Multilevel Hyperparameters:
# ~effect_id (Number of levels: 2361) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     1.00      0.00     1.00     1.00   NA       NA       NA
# 
# Regression Coefficients:
#           Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept    -0.32      0.09    -0.51    -0.15 1.00     1503     2737
# 
# Further Distributional Parameters:
#       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sigma     0.97      0.02     0.92     1.01 1.00     3047     4199
```

##### glmmTMB
```{r}
#| eval: false

dat_Roger$effect_id <- factor(dat_Roger$effect_id)
dat_Roger$site_id <- factor(dat_Roger$site_id)

VCV <- diag(dat_Roger$var_Hedges, nrow = nrow(dat_Roger)) 
rownames(VCV)<- colnames(VCV)<- dat_Roger$effect_id
VCV[1:5, 1:5]

dat_Roger$pos <- numFactor(dat_Roger$x_km, dat_Roger$y_km)

system.time(
  sp_eg1_exp_tmb <- glmmTMB(d_Hedges ~ 1 
                   + equalto(0+effect_id|const, VCV)
                   # + (1|site_id)
                   + exp(pos+0|const),
                   data = dat_Roger, 
                   REML=TRUE)
  )
 #   user  system elapsed 
 # 45.570   1.369  47.530 

summary(sp_eg1_exp_tmb)　
# Family: gaussian  ( identity )
# Formula:          d_Hedges ~ 1 + equalto(0 + effect_id | const, VCV) + exp(pos +      0 | const)
# Data: dat_Roger
# 
#       AIC       BIC    logLik -2*log(L)  df.resid 
#    9088.5    9111.5   -4540.2    9080.5      2357 
# 
# Random effects:
# 
# Conditional model:
#  Groups   Name                                     Variance  Std.Dev. Corr                                            
#  const    effect_id1                                0.202208 0.44968                                                  
#           effect_id10                               0.166676 0.40826  0.00                                            
#           effect_id100                              0.225770 0.47515  0.00 0.00                                       
#           effect_id1000                             0.197994 0.44497  0.00 0.00 0.00                                  
#           effect_id1001                             3.676968 1.91754  0.00 0.00 0.00 0.00                             
#           effect_id1002                             0.315616 0.56180  0.00 0.00 0.00 0.00 0.00                        
#           effect_id1003                             0.305460 0.55268  0.00 0.00 0.00 0.00 0.00 0.00                   
#           effect_id1004                             0.313426 0.55984  0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           effect_id1005                             0.413540 0.64307  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           effect_id1006                             0.325726 0.57072  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...                                      ...       ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  const.1  pos(16262.6644099893,-5204.51951303405)   1.866150 1.36607                                                  
#           pos(16280.4755285163,-5116.14629455727)   1.866150 1.36607  0.00                                            
#           pos(-7965.96086752978,-5019.4733863405)   1.866150 1.36607  0.00 0.00                                       
#           pos(-7959.34359171906,-4693.06364429579)  1.866150 1.36607  0.00 0.00 0.00                                  
#           pos(-7960.45678662699,-4691.63535918108)  1.866150 1.36607  0.00 0.00 0.00 0.00                             
#           pos(-7096.61753807119,-4685.92422115392)  1.866150 1.36607  0.00 0.00 0.00 0.00 0.00                        
#           pos(-7992.73943895704,-4607.717759142)    1.866150 1.36607  0.00 0.00 0.00 0.00 0.00 0.00                   
#           pos(-7903.68384632242,-4579.4258128701)   1.866150 1.36607  0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           pos(16237.0609271069,-4578.0132445546)    1.866150 1.36607  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           pos(16304.594788628,-4572.36489569299)    1.866150 1.36607  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...                                      ...       ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  Residual                                           1.864305 1.36540                                                  
# Number of obs: 2361, groups:  const, 1
# 
# Dispersion estimate for gaussian family (sigma^2): 1.86 
# 
# Conditional model:
#             Estimate Std. Error z value Pr(>|z|)    
# (Intercept) -0.40936    0.08139  -5.029 4.92e-07 ***

head(confint(sp_eg1_exp_tmb), 10) 
#                                                               2.5 %     97.5 %   Estimate
# (Intercept)                                              -0.5688798 -0.2498315 -0.4093556
# Std.Dev.pos(16262.6644099893,-5204.51951303405)|const.1   1.2417049  1.5028930  1.3660708
# Std.Dev.pos(16280.4755285163,-5116.14629455727)|const.1   1.2417049  1.5028930  1.3660708
# Std.Dev.pos(-7965.96086752978,-5019.4733863405)|const.1   1.2417049  1.5028930  1.3660708
# Std.Dev.pos(-7959.34359171906,-4693.06364429579)|const.1  1.2417049  1.5028930  1.3660708
# Std.Dev.pos(-7960.45678662699,-4691.63535918108)|const.1  1.2417049  1.5028930  1.3660708
# Std.Dev.pos(-7096.61753807119,-4685.92422115392)|const.1  1.2417049  1.5028930  1.3660708
# Std.Dev.pos(-7992.73943895704,-4607.717759142)|const.1    1.2417049  1.5028930  1.3660708
# Std.Dev.pos(-7903.68384632242,-4579.4258128701)|const.1   1.2417049  1.5028930  1.3660708
# Std.Dev.pos(16237.0609271069,-4578.0132445546)|const.1    1.2417049  1.5028930  1.3660708

sigma(sp_eg1_exp_tmb) 
# 1.365396
sp_eg1_exp_tmb_varcor <- VarCorr(sp_eg1_exp_tmb)$cond
exp(sp_eg1_exp_tmb$fit$par)
#   betadisp (efect_id)     theta (tau/tau^2)      theta (rho)
#   1.3653956               1.3660708 (1.866149)   0.2476221 

```

#### Gaussian process

##### metafor
```{r}

```
##### brms
```{r}

```

### 2. Maire et al. (2019)
We use a dataset from the `metadat` package as our second illustrative example. In the original study, spatial correlations among sampling stations were modelled using station-level coordinates along river networks. 
However, precise latitude and longitude information for individual stations is not available in the published dataset; only the locations of study sites can be geo-referenced. 
We therefore used site-level coordinates as proxies for station-level positions. Although this approximation allows us to compare the implementation of spatial meta-analytic models across `metafor`, `brms`, and `glmmTMB`, the resulting estimates are not directly comparable to those reported by Maire et al. (2019), where station-level spatial random effects were explicitly modelled.

```{r}
#| eval: false

dat <- dat.maire2019$dat
dmat <- dat.maire2019$dmat
dmat[1:35,1:35]

sites <- tribble(
  ~site_id, ~site_name,   ~lat,    ~lon,
  "site1",  "Belleville",  47.5,    2.65,
  "site2",  "Bugey",       45.8,    5.27,
  "site3",  "Chinon",      47.2,    0.18,
  "site4",  "Chooz",       50.1,    4.80,
  "site5",  "Civaux",      46.4,    0.65,
  "site6",  "Cruas",       44.6,    4.76,
  "site7",  "Dampierre",   47.7,    2.51,
  "site8",  "Nogent",      48.5,    3.52,
  "site9",  "St Alban",    45.3,    4.75,
  "site10", "St Laurent",  47.7,    2.53,
  "site11", "Tricastin",   44.3,    4.73
)


dat_maire2019 <- dat %>%
  left_join(sites, by = c("site" = "site_id")) |> 
  mutate(id = row_number())

dat_maire2019$const <- 1 # add constant for spatial models

# project to a planar coordinate system and convert to kilometers
dat_sf <- st_as_sf(dat_maire2019, coords = c("lon", "lat"), crs = 4326)
dat_sf_proj <- st_transform(dat_sf, crs = 3857)

coords_m <- st_coordinates(dat_sf_proj)
dat_maire2019$x_km <- coords_m[,1] / 1000
dat_maire2019$y_km <- coords_m[,2] / 1000

coords_km <- cbind(dat_maire2019$x_km, dat_maire2019$y_km)
dist_matrix_euclid <- as.matrix(dist(coords_km))

# row and column names
rownames(dist_matrix_euclid) <- dat_maire2019$id
colnames(dist_matrix_euclid) <- dat_maire2019$id

head(dat_maire2019)
```

#### Exponential kernel
##### metafor
```{r}
#| eval: false

ma_sp2_exp <- rma.mv(s1, vars1, 
             random = list(
                       ~ 1|id,
                       ~ 1|site,
                       ~ x_km + y_km | const), 
             struct = "SPEXP", 
             data = dat_maire2019,
             test = "t",
             method = "REML",
             sparse = TRUE,
             verbose = TRUE
             )

summary(ma_sp2_exp)
# Multivariate Meta-Analysis Model (k = 35; method: REML)
# 
# logLik   Deviance        AIC        BIC       AICc   
# -199.0469   398.0939   408.0939   415.7257   410.2367   
# 
# Variance Components:
#   
#   estim     sqrt  nlvls  fixed  factor 
# sigma^2.1   819.5017  28.6269     35     no      id 
# sigma^2.2  1653.5229  40.6635     11     no    site 
# 
# outer factor: const        (nlvls = 1)
# inner term:   ~x_km + y_km (nlvls = 11)
# 
# estim     sqrt  fixed 
# tau^2      8548.0233  92.4555     no 
# rho          97.3090              no 
# 
# Test for Heterogeneity:
#   Q(df = 34) = 191.4837, p-val < .0001
# 
# Model Results:
#   
#   estimate       se    zval    pval     ci.lb     ci.ub      
# 190.8255  40.9187  4.6635  <.0001  110.6263  271.0246  ***

confint(ma_sp2_exp)
# estimate  ci.lb     ci.ub 
# sigma^2.1 819.5017 0.0000 4240.3900 
# sigma.1    28.6269 0.0000   65.1183 
# 
# estimate  ci.lb      ci.ub 
# sigma^2.2 1653.5229 0.0000 21698.5229 
# sigma.2     40.6635 0.0000   147.3042 
# 
# estimate  ci.lb        ci.ub 
# tau^2 8548.0233 0.0000 >854802.3263 
# tau     92.4555 0.0000    >924.5552 
# 
# estimate   ci.lb     ci.ub 
# rho  97.3090 <9.7309 >973.0899 

```

```{r}

```

##### brms
```{r}
#| eval: false

dat_maire2019$id <- as.character(dat_maire2019$id)
vcv <- diag(dat_maire2019$vars1)

rownames(vcv) <- colnames(vcv) <- dat_maire2019$id

fit_sp_eg2_exp_brms <- bf(s1 ~ 1 +
                 # (1|site) + # this is 
                 (1|gr(id, cov = vcv)) + # this is m (sampling error)
                gp(x_km, y_km, 
                   cov = "exponential",
                   scale = FALSE)
                )

# generate default priors
prior <- default_prior(fit_sp_eg2_exp_brms, 
                           data = dat_maire2019, 
                           data2 = list(vcv = vcv),
                           family = gaussian())
prior$prior[5] = "constant(1)" # meta-analysis assumes sampling variance is known so fixing this to 1
prior

# fitting model
sp_eg2_exp_brms <- brm(
  formula = fit_sp_eg2_exp_brms,
  data = dat_maire2019,
  data2 = list(vcv=vcv),
  chains = 2,
  threads = threading(threads_per_chain),
  iter = 15000,
  warmup = 12000,
  prior = prior,
  control = list(adapt_delta=0.95, max_treedepth=15)
)
summary(sp_eg2_exp_brms)
```

##### glmmTMB
```{r}
#| eval: false

dat_maire2019$id <- factor(dat_maire2019$id)
VCV <- diag(dat_maire2019$vars1, nrow = nrow(dat_maire2019)) 
rownames(VCV)<- colnames(VCV)<- dat_maire2019$id
VCV[1:5, 1:5]

dat_maire2019$pos <- numFactor(dat_maire2019$x_km, dat_maire2019$y_km)

tmb_sp2_exp <- glmmTMB(
  s1 ~ 1 +
    equalto(0 + id | const, VCV) + 
    (1 | site) +
    exp(pos + 0 | const),     
  data = dat_maire2019,
  REML = TRUE
  )


head(confint(tmb_sp2_exp), 10)

sigma(tmb_sp2_exp) # for id

tmb_sp2_exp_varcor <- VarCorr(tmb_sp2_exp)$cond
exp(tmb_sp2_exp$fit$par[[4]]) # for spatial range parameter rho


```

#### Gaussian process

##### metafor
```{r}
#| eval: false

ma_sp2_gau <- rma.mv(s1, vars1, 
                          random = list(
                            ~ 1|id,
                            ~ 1|site,
                            ~ x_km + y_km | const), 
                          struct = "SPGAU", 
                          data = dat_maire2019,
                          sparse = TRUE,
                          verbose = TRUE
)

summary(ma_sp2_gau)
# Multivariate Meta-Analysis Model (k = 35; method: REML)
# 
# logLik   Deviance        AIC        BIC       AICc   
# -199.1047   398.2094   408.2094   415.8412   410.3522   
# 
# Variance Components:
#   
#   estim     sqrt  nlvls  fixed  factor 
# sigma^2.1   805.9835  28.3898     35     no      id 
# sigma^2.2  3557.3247  59.6433     11     no    site 
# 
# outer factor: const        (nlvls = 1)
# inner term:   ~x_km + y_km (nlvls = 11)
# 
# estim     sqrt  fixed 
# tau^2      6637.4779  81.4707     no 
# rho         143.9412              no 
# 
# Test for Heterogeneity:
#   Q(df = 34) = 191.4837, p-val < .0001
# 
# Model Results:
#   
#   estimate       se    zval    pval     ci.lb     ci.ub      
# 186.2795  40.6944  4.5775  <.0001  106.5199  266.0392  *** 

confint(ma_sp2_gau)

```

##### brms
```{r}


```

##### glmmTMB
```{r}
#| eval: false

tmb_sp2_gau <- glmmTMB(
  s1 ~ 1 +
    equalto(0 + id | const, VCV) + 
    (1 | site) +
    gau(pos + 0 | const),     
  data = dat_maire2019,
  REML = TRUE
)

head(confint(tmb_sp2_gau), 10)

tmb_sp2_gau_varcor <- VarCorr(tmb_sp2_gau)$cond
exp(tmb_sp2_gau$fit$par[[4]]) # for spatial range parameter rho
```

:::

# Software and package versions

```{r}
#| echo: false

sessionInfo()

```


# References
-   Maire, A., Thierry, E., Viechtbauer, W., & Daufresne, M. (2019). Poleward shift in large-river fish communities detected with a novel meta-analysis framework. Freshwater Biology, 64(6), 1143–1156. <https://doi.org/10.1111/fwb.13291>
-   Grau‐Andrés, R., Moreira, B., & Pausas, J. G. (2024). Global plant responses to intensified fire regimes. Global Ecology and Biogeography, 33(8), e13858. <https://doi.org/10.1111/geb.13858>
-    Viechtbauer, W., White, T., Noble, D., Senior, A., & Hamilton, W. K. (2025). metadat: Meta-analysis datasets (Version 1.5-2) [R package]. <https://github.com/wviechtb/metadat>
-    Rios Moura, R., Oliveira Gonzaga, M., Silva Pinto, N., Vasconcellos-Neto, J., & Requena, G. S. (2021). Assortative mating in space and time: Patterns and biases. Ecology Letters, 24(5), 1089–1102. <https://doi.org/10.1111/ele.13690>
-   Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, and A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491:444-448. <https://doi.org/10.1038/nature11631>


