---
title: "Phylo spatial meta-analysis"
date: "`r Sys.Date()`"
# author: " "
format: 
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-float: true
    toc-title: "**CONTENTS**"
    output-file: "AM_worked_example.html"
    embed-resources: true
    # code-fold: true
    code-tools: 
      source: false
    number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "12"
    mainfont: "Roboto"
    monofont: "Roboto"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: united
    code-line-numbers: false
    grid:
      margin-width: 180px
crossref: 
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
  # cache: true
---

# Setup
```{r}
install.packages("pacman")
pacman::p_load(brms, metafor, metadat, tidyverse, data.table, crayon, here, ape, patchwork,
               dplyr, tidyr,MCMCglmm, bayesplot, tidybayes, orchaRd,purrr, stringr, readr,
               lubridate, magrittr, janitor, glmmTMB, rotl, sf)
```
# Phylogenetic meta-analysis

## How to make phylogenetic correlation matrix?
```{r}

```

### Using a specific phylogenetic tree {#using-a-specific-phylogenetic-tree}
```{r}

```

### Using the `rotl` package (Open Tree of Life) {#using-the-rotl-package-open-tree-of-life}
```{r}

```

## Examples

::: panel-tabset
### 1
```{r}
dat_moura2021 <- dat.moura2021$dat
tree <- dat.moura2021$tree

# calculate r-to-z transformed correlations and corresponding sampling variances
dat_moura2021 <- escalc(measure="ZCOR", ri=ri, ni=ni, data=dat_moura2021)

# turn tree into an ultrametric one
tree <- compute.brlen(tree)

# compute phylogenetic correlation matrix
A <- vcv(tree, corr = TRUE)

# make copy of the species.id variable
dat_moura2021$species.id.phy <- dat_moura2021$species.id

```

#### Brownian motion

##### metafor

##### brms
```{r}

```

##### glmmTMB
```{r}

```

#### Ornstein-Uhlenbeck

##### metafor
```{r}

```

### 2

#### Brownian motion

##### metafor
```{r}

```

##### brms
```{r}

```

##### glmmTMB
```{r}

```

#### Ornstein-Uhlenbeck

##### metafor
```{r}

```

:::

# Spatial meta-analysis

## How to make spatial correlation matrix?

## Examples

::: panel-tabset
### 1. Grau-Andrés et al. (2024)

```{r}

dat_Roger <- read.csv(here("data", "examples", "Roger_etal_2024", "Roger_etal_2024.csv"))
head(dat_Roger)

dat_Roger$const <- 1 # add constant for spatial models
dat_Roger$effect_id <- seq_len(nrow(dat_Roger))

dat_Roger <- dat_Roger %>%
  filter(!is.na(longitude))
coords2 <- cbind(dat_Roger$longitude, dat_Roger$latitude)

# project to a planar coordinate system and convert to kilometers
dat_sf2 <- st_as_sf(dat_Roger, coords = c("longitude", "latitude"), crs = 4326)
dat_sf2_proj <- st_transform(dat_sf2, crs = 3857)

coords_m <- st_coordinates(dat_sf2_proj)
dat_Roger$x_km <- coords_m[,1] / 1000
dat_Roger$y_km <- coords_m[,2] / 1000

coords_km <- cbind(dat_Roger$x_km, dat_Roger$y_km)
dist_matrix_euclid2 <- as.matrix(dist(coords_km))

# row and column names
rownames(dist_matrix_euclid2) <- dat_Roger$effect_id
colnames(dist_matrix_euclid2) <- dat_Roger$effect_id

dat_Roger <- dat_Roger %>%
  group_by(latitude, longitude) %>%
  mutate(site_id = cur_group_id()) %>%
  ungroup()
names(dat_Roger)
```

#### Exponential kernel

##### metafor
```{r}
#| eval: false
system.time(sp_eg1_exp_mf <- rma.mv(d_Hedges, var_Hedges, 
                              random = list(
                                ~ 1|effect_id,
                                ~ 1|site_id,
                                ~ x_km + y_km |const
                              ), 
                              struct = "SPEXP", 
                              data = dat_Roger,
                              sparse = TRUE,
                              verbose = TRUE,
                              method = "REML",
                              test = "t"
                              )
            )
#     user   system  elapsed 
#   969.641   24.550 1018.565 

summary(sp_eg1_exp_mf)
# Multivariate Meta-Analysis Model (k = 2361; method: REML)
# 
#     logLik    Deviance         AIC         BIC        AICc   
# -4031.5291   8063.0583   8073.0583   8101.8903   8073.0837   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed     factor 
# sigma^2.1  0.7940  0.8911   2361     no  effect_id 
# sigma^2.2  0.9288  0.9637    383     no    site_id 
# 
# outer factor: const        (nlvls = 1)
# inner term:   ~x_km + y_km (nlvls = 383)
# 
#               estim    sqrt  fixed 
# tau^2        0.2989  0.5467     no 
# rho        188.1869             no 
# 
# Test for Heterogeneity:
# Q(df = 2360) = 21273.5149, p-val < .0001
# 
# Model Results:
# 
# estimate      se     tval    df    pval    ci.lb    ci.ub      
#  -0.3347  0.0765  -4.3727  2360  <.0001  -0.4847  -0.1846  *** 
```
##### brms
```{r}

```

```{r}
dat_Roger$effect_id <- as.character(dat_Roger$effect_id)
vcv <- diag(dat_Roger$var_Hedges)

rownames(vcv) <- colnames(vcv) <- dat_Roger$effect_id

fit_sp_eg1_exp_brms <- bf(d_Hedges ~ 1 +
                 (1|site_id) + # this is 
                 (1|gr(effect_id, cov = vcv)) + # this is m (sampling error)
                gp(x_km, y_km, 
                   cov = "exponential",
                   scale = FALSE)
                )

# generate default priors
prior <- default_prior(fit_sp_eg1_exp_brms, 
                           data = dat_Roger, 
                           data2 = list(vcv = vcv),
                           family = gaussian())
prior$prior[5] = "constant(1)" # meta-analysis assumes sampling variance is known so fixing this to 1
prior

# fitting model
sp_eg1_exp_brms <- brm(
  formula = fit_eg4_v1,
  data = dat_Roger,
  data2 = list(vcv=vcv),
  chains = 2,
  iter = 6000,
  warmup = 3000,
  prior = prior,
  control = list(adapt_delta=0.95, max_treedepth=15)
)
summary(sp_eg1_exp_brms)
# Family: gaussian 
#   Links: mu = identity 
# Formula: d_Hedges ~ 1 + (1 | study_id) + (1 | gr(effect_id, cov = vcv)) + gp(x_km, y_km, cov = "exponential", scale = FALSE) 
#    Data: dat_Roger (Number of observations: 2361) 
#   Draws: 2 chains, each with iter = 6000; warmup = 3000; thin = 1;
#          total post-warmup draws = 6000
# 
# Gaussian Process Hyperparameters:
#                    Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sdgp(gpx_kmy_km)       0.20      0.12     0.01     0.47 1.00      344     1056
# lscale(gpx_kmy_km)  5258.38  31234.81   440.25 24312.38 1.00      822     1513
# 
# Multilevel Hyperparameters:
# ~effect_id (Number of levels: 2361) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     1.00      0.00     1.00     1.00   NA       NA       NA
# 
# ~study_id (Number of levels: 393) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     1.10      0.06     0.99     1.21 1.00     1331     2059
# 
# Regression Coefficients:
#           Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept    -0.34      0.10    -0.53    -0.12 1.00     1794     2030
# 
# Further Distributional Parameters:
#       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sigma     0.88      0.02     0.84     0.92 1.00     2663     3946
```

##### glmmTMB
```{r}
#| eval: false

dat_Roger$effect_id <- factor(dat_Roger$effect_id)
dat_Roger$site_id <- factor(dat_Roger$site_id)

VCV <- diag(dat_Roger$var_Hedges, nrow = nrow(dat_Roger)) 
rownames(VCV)<- colnames(VCV)<- dat_Roger$effect_id
VCV[1:5, 1:5]

dat_Roger$pos <- numFactor(dat_Roger$x_km, dat_Roger$y_km)

system.time(
  sp_eg1_exp_tmb <- glmmTMB(d_Hedges ~ 1 
                   + equalto(0+effect_id|const, VCV)
                   + (1|site_id)
                   + exp(pos+0|const),
                   data = dat_Roger, 
                   REML=TRUE)
  )
 #   user  system elapsed 
 # 45.570   1.369  47.530 

summary(sp_eg1_exp_tmb)　
# Family: gaussian  ( identity )
# Formula:          d_Hedges ~ 1 + equalto(0 + effect_id | const, VCV) + (1 | site_id) +      exp(pos + 0 | const)
# Data: dat_Roger
# 
#       AIC       BIC    logLik -2*log(L)  df.resid 
#    9090.5    9119.3   -4540.2    9080.5      2356 
# 
# Random effects:
# 
# Conditional model:
#  Groups   Name                                     Variance  Std.Dev. Corr                                            
#  const    effect_id1                                0.202208 0.44968                                                  
#           effect_id10                               0.166676 0.40826  0.00                                            
#           effect_id100                              0.225770 0.47515  0.00 0.00                                       
#           effect_id1000                             0.197994 0.44497  0.00 0.00 0.00                                  
#           effect_id1001                             3.676968 1.91754  0.00 0.00 0.00 0.00                             
#           effect_id1002                             0.315616 0.56180  0.00 0.00 0.00 0.00 0.00                        
#           effect_id1003                             0.305460 0.55268  0.00 0.00 0.00 0.00 0.00 0.00                   
#           effect_id1004                             0.313426 0.55984  0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           effect_id1005                             0.413540 0.64307  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           effect_id1006                             0.325726 0.57072  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...                                      ...       ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  site_id  (Intercept)                               0.123984 0.35211                                                  
#  const.1  pos(16262.6644099893,-5204.51951303405)   1.741641 1.31971                                                  
#           pos(16280.4755285163,-5116.14629455727)   1.741641 1.31971  0.00                                            
#           pos(-7965.96086752978,-5019.4733863405)   1.741641 1.31971  0.00 0.00                                       
#           pos(-7959.34359171906,-4693.06364429579)  1.741641 1.31971  0.00 0.00 0.00                                  
#           pos(-7960.45678662699,-4691.63535918108)  1.741641 1.31971  0.00 0.00 0.00 0.00                             
#           pos(-7096.61753807119,-4685.92422115392)  1.741641 1.31971  0.00 0.00 0.00 0.00 0.00                        
#           pos(-7992.73943895704,-4607.717759142)    1.741641 1.31971  0.00 0.00 0.00 0.00 0.00 0.00                   
#           pos(-7903.68384632242,-4579.4258128701)   1.741641 1.31971  0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           pos(16237.0609271069,-4578.0132445546)    1.741641 1.31971  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           pos(16304.594788628,-4572.36489569299)    1.741641 1.31971  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...                                      ...       ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  Residual                                           1.864281 1.36539                                                  
# Number of obs: 2361, groups:  const, 1; site_id, 383
# 
# Dispersion estimate for gaussian family (sigma^2): 1.86 
# 
# Conditional model:
#             Estimate Std. Error z value Pr(>|z|)    
# (Intercept) -0.40925    0.08138  -5.029 4.93e-07 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
head(confint(sp_eg1_exp_tmb), 10) 

#                                                                  2.5 %       97.5 %   Estimate
# (Intercept)                                              -5.687455e-01   -0.2497545 -0.4092500
# Std.Dev.(Intercept)|site_id                               1.590721e-05 7794.2085876  0.3521138
# Std.Dev.pos(16262.6644099893,-5204.51951303405)|const.1   6.410538e-01    2.7168414  1.3197126
# Std.Dev.pos(16280.4755285163,-5116.14629455727)|const.1   6.410538e-01    2.7168414  1.3197126
# Std.Dev.pos(-7965.96086752978,-5019.4733863405)|const.1   6.410538e-01    2.7168414  1.3197126
# Std.Dev.pos(-7959.34359171906,-4693.06364429579)|const.1  6.410538e-01    2.7168414  1.3197126
# Std.Dev.pos(-7960.45678662699,-4691.63535918108)|const.1  6.410538e-01    2.7168414  1.3197126
# Std.Dev.pos(-7096.61753807119,-4685.92422115392)|const.1  6.410538e-01    2.7168414  1.3197126
# Std.Dev.pos(-7992.73943895704,-4607.717759142)|const.1    6.410538e-01    2.7168414  1.3197126
# Std.Dev.pos(-7903.68384632242,-4579.4258128701)|const.1   6.410538e-01    2.7168414  1.3197126


sigma(sp_eg1_exp_tmb) 
# 1.365387
sp_eg1_exp_tmb_varcor <- VarCorr(sp_eg1_exp_tmb)$cond
exp(sp_eg1_exp_tmb$fit$par[[2]])
exp(sp_eg1_exp_tmb$fit$par)
#  betadisp     theta     theta     theta 
# 1.3653868 0.3521138 1.3197126 0.2527374
```

##### summarise results
```{r}
meta <- data.frame(model = "metafor", 
                       logLik = logLik(sp_eg1_exp_mf),
                       est = sp_eg1_exp_mf$b[[1]], 
                       se = sp_eg1_exp_mf$se[[1]], 
                       sigma2.site = sp_eg1_exp_mf$sigma2[2], ## among study variance
                       sigma2.effect = sp_eg1_exp_mf$sigma2[1], ## within study variance
                       tau2 = sp_eg1_exp_mf$tau2,
                       rho = sp_eg1_exp_mf$rho
                   )


tmb <- data.frame(model = "glmmTMB",
                   logLik = logLik(sp_eg1_exp_tmb)[1],
                   est = unlist(fixef(sp_eg1_exp_tmb))[[1]], #overall mean
                   se = as.numeric(sqrt(vcov(sp_eg1_exp_tmb)[[1]])), #overall mean SE
                   sigma2.site = (sp_eg1_exp_tmb_varcor$site_id[1]), ## among site variance estimate
                   sigma2.effect = sigma(sp_eg1_exp_tmb), ## within study variance estimate -> effect id
                   tau2 = exp(sp_eg1_exp_tmb$fit$par[3]), ##spatial variance estimate
                   rho = exp(sp_eg1_exp_tmb$fit$par[4]) ## rho
                  )

summ <- posterior_summary(
  sp_eg1_exp_brms,
  variable = c(
    "b_Intercept",
    "sigma",
    "sd_study_id__Intercept",
    "sdgp_gpx_kmy_km",
    "lscale_gpx_kmy_km"
  )
)

brms <- data.frame(
  model = "brms",
  logLik = NA_real_, 
  est = summ["b_Intercept","Estimate"],
  se  = summ["b_Intercept","Est.Error"],
  sigma2.site = (summ["sd_study_id__Intercept","Estimate"])^2,
  sigma2.effect = (summ["sigma","Estimate"])^2,
  tau2 = (summ["sdgp_gpx_kmy_km","Estimate"])^2,
  rho = summ["lscale_gpx_kmy_km","Estimate"]
)

output <- rbind(meta, tmb, brms)

knitr::kable(output)
```


#### Gaussian process

##### metafor
```{r}

```
##### brms
```{r}

```

### 2. Maire et al. (2019)
We use a dataset from the `metadat` package as our second illustrative example. In the original study, spatial correlations among sampling stations were modelled using station-level coordinates along river networks. 
However, precise latitude and longitude information for individual stations is not available in the published dataset; only the locations of study sites can be geo-referenced. 
We therefore used site-level coordinates as proxies for station-level positions. Although this approximation allows us to compare the implementation of spatial meta-analytic models across `metafor`, `brms`, and `glmmTMB`, the resulting estimates are not directly comparable to those reported by Maire et al. (2019), where station-level spatial random effects were explicitly modelled.

```{r}
dat <- dat.maire2019$dat
dmat <- dat.maire2019$dmat
dmat[1:35,1:35]

sites <- tribble(
  ~site_id, ~site_name,   ~lat,    ~lon,
  "site1",  "Belleville",  47.5,    2.65,
  "site2",  "Bugey",       45.8,    5.27,
  "site3",  "Chinon",      47.2,    0.18,
  "site4",  "Chooz",       50.1,    4.80,
  "site5",  "Civaux",      46.4,    0.65,
  "site6",  "Cruas",       44.6,    4.76,
  "site7",  "Dampierre",   47.7,    2.51,
  "site8",  "Nogent",      48.5,    3.52,
  "site9",  "St Alban",    45.3,    4.75,
  "site10", "St Laurent",  47.7,    2.53,
  "site11", "Tricastin",   44.3,    4.73
)


dat_maire2019 <- dat %>%
  left_join(sites, by = c("site" = "site_id")) |> 
  mutate(id = row_number())

dat_maire2019$const <- 1 # add constant for spatial models

# project to a planar coordinate system and convert to kilometers
dat_sf <- st_as_sf(dat_maire2019, coords = c("lon", "lat"), crs = 4326)
dat_sf_proj <- st_transform(dat_sf, crs = 3857)

coords_m <- st_coordinates(dat_sf_proj)
dat_maire2019$x_km <- coords_m[,1] / 1000
dat_maire2019$y_km <- coords_m[,2] / 1000

coords_km <- cbind(dat_maire2019$x_km, dat_maire2019$y_km)
dist_matrix_euclid <- as.matrix(dist(coords_km))

# row and column names
rownames(dist_matrix_euclid) <- dat_maire2019$id
colnames(dist_matrix_euclid) <- dat_maire2019$id

head(dat_maire2019)
```

#### Exponential kernel
##### metafor
```{r}
#| eval: false

ma_sp2_exp <- rma.mv(s1, vars1, 
             random = list(
                       ~ 1|id,
                       ~ 1|site,
                       ~ x_km + y_km | const), 
             struct = "SPEXP", 
             data = dat_maire2019,
             test = "t",
             method = "REML",
             sparse = TRUE,
             verbose = TRUE
             )

summary(ma_sp2_exp)
# Multivariate Meta-Analysis Model (k = 35; method: REML)
# 
# logLik   Deviance        AIC        BIC       AICc   
# -199.0469   398.0939   408.0939   415.7257   410.2367   
# 
# Variance Components:
#   
#   estim     sqrt  nlvls  fixed  factor 
# sigma^2.1   819.5017  28.6269     35     no      id 
# sigma^2.2  1653.5229  40.6635     11     no    site 
# 
# outer factor: const        (nlvls = 1)
# inner term:   ~x_km + y_km (nlvls = 11)
# 
# estim     sqrt  fixed 
# tau^2      8548.0233  92.4555     no 
# rho          97.3090              no 
# 
# Test for Heterogeneity:
#   Q(df = 34) = 191.4837, p-val < .0001
# 
# Model Results:
#   
#   estimate       se    zval    pval     ci.lb     ci.ub      
# 190.8255  40.9187  4.6635  <.0001  110.6263  271.0246  ***

confint(ma_sp2_exp)
# estimate  ci.lb     ci.ub 
# sigma^2.1 819.5017 0.0000 4240.3900 
# sigma.1    28.6269 0.0000   65.1183 
# 
# estimate  ci.lb      ci.ub 
# sigma^2.2 1653.5229 0.0000 21698.5229 
# sigma.2     40.6635 0.0000   147.3042 
# 
# estimate  ci.lb        ci.ub 
# tau^2 8548.0233 0.0000 >854802.3263 
# tau     92.4555 0.0000    >924.5552 
# 
# estimate   ci.lb     ci.ub 
# rho  97.3090 <9.7309 >973.0899 

```

```{r}

```

##### brms
```{r}

```

##### glmmTMB
```{r}

dat_maire2019$id <- factor(dat_maire2019$id)
VCV <- diag(dat_maire2019$vars1, nrow = nrow(dat_maire2019)) 
rownames(VCV)<- colnames(VCV)<- dat_maire2019$id
VCV[1:5, 1:5]

dat_maire2019$pos <- numFactor(dat_maire2019$x_km, dat_maire2019$y_km)

tmb_sp2_exp <- glmmTMB(
  s1 ~ 1 +
    equalto(0 + id | const, VCV) + 
    (1 | site) +
    exp(pos + 0 | const),     
  data = dat_maire2019,
  REML = TRUE
  )


head(confint(tmb_sp2_exp), 10)

sigma(tmb_sp2_exp) # for id

tmb_sp2_exp_varcor <- VarCorr(tmb_sp2_exp)$cond
exp(tmb_sp2_exp$fit$par[[4]]) # for spatial range parameter rho


```

#### Gaussian process

##### metafor
```{r}

ma_sp2_gau <- rma.mv(s1, vars1, 
                          random = list(
                            ~ 1|id,
                            ~ 1|site,
                            ~ x_km + y_km | const), 
                          struct = "SPGAU", 
                          data = dat_maire2019,
                          sparse = TRUE,
                          verbose = TRUE
)

summary(ma_sp2_gau)
# Multivariate Meta-Analysis Model (k = 35; method: REML)
# 
# logLik   Deviance        AIC        BIC       AICc   
# -199.1047   398.2094   408.2094   415.8412   410.3522   
# 
# Variance Components:
#   
#   estim     sqrt  nlvls  fixed  factor 
# sigma^2.1   805.9835  28.3898     35     no      id 
# sigma^2.2  3557.3247  59.6433     11     no    site 
# 
# outer factor: const        (nlvls = 1)
# inner term:   ~x_km + y_km (nlvls = 11)
# 
# estim     sqrt  fixed 
# tau^2      6637.4779  81.4707     no 
# rho         143.9412              no 
# 
# Test for Heterogeneity:
#   Q(df = 34) = 191.4837, p-val < .0001
# 
# Model Results:
#   
#   estimate       se    zval    pval     ci.lb     ci.ub      
# 186.2795  40.6944  4.5775  <.0001  106.5199  266.0392  *** 

confint(ma_sp2_gau)

```

##### brms
```{r}


```

##### glmmTMB
```{r}

tmb_sp2_gau <- glmmTMB(
  s1 ~ 1 +
    equalto(0 + id | const, VCV) + 
    (1 | site) +
    gau(pos + 0 | const),     
  data = dat_maire2019,
  REML = TRUE
)

head(confint(tmb_sp2_gau), 10)

tmb_sp2_gau_varcor <- VarCorr(tmb_sp2_gau)$cond
exp(tmb_sp2_gau$fit$par[[4]]) # for spatial range parameter rho
```

:::

# Software and package versions

```{r}
#| echo: false

sessionInfo()

```


# References
-   Maire, A., Thierry, E., Viechtbauer, W., & Daufresne, M. (2019). Poleward shift in large-river fish communities detected with a novel meta-analysis framework. Freshwater Biology, 64(6), 1143–1156. <https://doi.org/10.1111/fwb.13291>
-   Grau‐Andrés, R., Moreira, B., & Pausas, J. G. (2024). Global plant responses to intensified fire regimes. Global Ecology and Biogeography, 33(8), e13858. <https://doi.org/10.1111/geb.13858>
-    Viechtbauer, W., White, T., Noble, D., Senior, A., & Hamilton, W. K. (2025). metadat: Meta-analysis datasets (Version 1.5-2) [R package]. <https://github.com/wviechtb/metadat>

-    Rios Moura, R., Oliveira Gonzaga, M., Silva Pinto, N., Vasconcellos-Neto, J., & Requena, G. S. (2021). Assortative mating in space and time: Patterns and biases. Ecology Letters, 24(5), 1089–1102. https://doi.org/10.1111/ele.13690


