---
title: "Phylo spatial meta-analysis"
date: "`r Sys.Date()`"
# author: " "
format: 
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-float: true
    toc-title: "**CONTENTS**"
    output-file: "AM_worked_example.html"
    embed-resources: true
    # code-fold: true
    code-tools: 
      source: false
    number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "12"
    mainfont: "Roboto"
    monofont: "Roboto"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: united
    code-line-numbers: false
    grid:
      margin-width: 180px
crossref: 
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
  # cache: true
---

# Setup
```{r}
install.packages("pacman")
pacman::p_load(brms, metafor, metadat, tidyverse, data.table, crayon, here, ape, patchwork,
               dplyr, tidyr,MCMCglmm, bayesplot, tidybayes, orchaRd,purrr, stringr, readr,
               lubridate, magrittr, janitor, glmmTMB, rotl, sf)

max_cores <- 10
num_chains <- 2
threads_per_chain <- floor(max_cores / num_chains)
options(mc.cores = num_chains) 
```

# Phylogenetic meta-analysis

## How to make phylogenetic correlation matrix?
```{r}

```

### Using a specific phylogenetic tree {#using-a-specific-phylogenetic-tree}
```{r}

```

### Using the `rotl` package (Open Tree of Life) {#using-the-rotl-package-open-tree-of-life}
```{r}

```

## Example

::: panel-tabset
###  Moura et al. (2021)
```{r}
dat_moura2021 <- dat.moura2021$dat
tree <- dat.moura2021$tree

# calculate r-to-z transformed correlations and corresponding sampling variances
dat_moura2021 <- escalc(measure="ZCOR", ri=ri, ni=ni, data=dat_moura2021)

# turn tree into an ultrametric one
tree <- compute.brlen(tree)

# compute phylogenetic correlation matrix
A <- vcv(tree, corr = TRUE)

# make copy of the species.id variable
dat_moura2021$species.id.phy <- dat_moura2021$species.id

```

#### Brownian motion

##### metafor
```{r}
#| echo: false

phylo_eg1_meta_BM <- readRDS(here("Rdata", "tutorial_v2", "moura2021_BM_metafor.rds"))

```

```{r}

system.time(
  phylo_eg1_meta_BM <- rma.mv(yi, vi,
         random = list(~ 1 | study.id, 
         ~ 1 | effect.size.id, 
         ~ 1 | species.id, 
         ~ 1 | species.id.phy),
         R = list(species.id.phy=A), 
         data = dat_moura2021,
         verbose = TRUE,
         sparse = TRUE,
         method = "REML"
         )
  )

summary(phylo_eg1_meta_BM)
# Multivariate Meta-Analysis Model (k = 1828; method: REML)

#    logLik   Deviance        AIC        BIC       AICc   
# -167.6726   335.3453   345.3453   372.8974   345.3782   

# Variance Components:

#             estim    sqrt  nlvls  fixed          factor    R 
# sigma^2.1  0.0192  0.1384    457     no        study.id   no 
# sigma^2.2  0.0145  0.1202   1828     no  effect.size.id   no 
# sigma^2.3  0.0557  0.2359    341     no      species.id   no 
# sigma^2.4  0.0512  0.2263    341     no  species.id.phy  yes 

# Test for Heterogeneity:
# Q(df = 1827) = 10743.8076, p-val < .0001

# Model Results:

# estimate      se    zval    pval   ci.lb   ci.ub     
#   0.3682  0.1300  2.8311  0.0046  0.1133  0.6230  ** 

# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

confint(phylo_eg1_meta_BM)
#           estimate  ci.lb  ci.ub 
# sigma^2.1   0.0192 0.0108 0.0325 
# sigma.1     0.1384 0.1038 0.1802 

#           estimate  ci.lb  ci.ub 
# sigma^2.2   0.0145 0.0121 0.0172 
# sigma.2     0.1202 0.1099 0.1311 

#           estimate  ci.lb  ci.ub 
# sigma^2.3   0.0557 0.0334 0.0788 
# sigma.3     0.2359 0.1827 0.2807 

#           estimate  ci.lb  ci.ub 
# sigma^2.4   0.0512 0.0179 0.1792 
# sigma.4     0.2263 0.1336 0.4233 

orchaRd::i2_ml(phylo_eg1_meta_BM)

phylo_heritability <- phylo_eg1_meta_BM$sigma2[4]^2 / (phylo_eg1_meta_BM$sigma2[4]^2 + phylo_eg1_meta_BM$sigma2[3]^2)
```

##### brms
```{r}
#| echo: false

phylo_eg1_brms <- readRDS(here("Rdata", "tutorial_v2", "moura2021_BM_brms2.rds"))

```

```{r}

fit_phylo_eg1_brms <- bf(yi | se(sqrt(vi)) ~ 1 +
                     (1 | study.id) + 
                     (1 | effect.size.id) + 
                     (1 | species.id) + 
                     (1 | gr(species.id.phy, cov = A))
                     )

prior_brms2 <- get_prior(
  formula = fit_brms2,
  data = dat_moura2021,
  data2 = list(A = A),
  family = gaussian()
  )

system.time(
  phylo_eg1_brms <- brm(
    formula = fit_phylo_eg1_brms,
    family = gaussian(),
    data = dat_moura2021,
    data2 = list(A = A),
    prior = prior_brms2,
    iter = 10000, 
    warmup = 7000,  
    chains = num_chains,
    backend = "cmdstanr",
    threads = threading(threads_per_chain), 
    control = list(adapt_delta = 0.95, max_treedepth = 15)
    )
  )

summary(phylo_eg1_brms)
#  Family: gaussian 
#   Links: mu = identity; sigma = identity 
# Formula: yi | se(sqrt(vi)) ~ 1 + spatially.pooled + (1 | study.id) + (1 | effect.size.id) + (1 | species.id) + (1 | gr(species.id.phy, cov = A)) 
#    Data: dat_moura2021 (Number of observations: 1828) 
#   Draws: 2 chains, each with iter = 10000; warmup = 7000; thin = 1;
#          total post-warmup draws = 6000

# Multilevel Hyperparameters:
# ~effect.size.id (Number of levels: 1828) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.12      0.01     0.11     0.13 1.00     1494     2868

# ~species.id (Number of levels: 341) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.23      0.03     0.17     0.28 1.00      535      624

# ~species.id.phy (Number of levels: 341) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.27      0.09     0.15     0.49 1.00      611      467

# ~study.id (Number of levels: 457) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     0.14      0.02     0.10     0.19 1.00      454      794

# Regression Coefficients:
#                     Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept               0.36      0.16     0.04     0.68 1.00     2050     2076
# spatially.pooledyes     0.04      0.03    -0.01     0.10 1.00     1800     2714

# Further Distributional Parameters:
#       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sigma     0.00      0.00     0.00     0.00   NA       NA       NA

```

##### glmmTMB
```{r}
phylo_eg1_tmb <- readRDS(here("Rdata", "tutorial_v2", "phylo_eg1_tmb1.rds"))
```

```{r}

A <- A[sort(rownames(A)), sort(rownames(A))]

dat_moura2021$g <- 1 

VCV <- diag(dat_moura2021$vi, nrow = nrow(dat_moura2021)) 
rownames(VCV)<- colnames(VCV)<- dat_moura2021$effect.size.id

dat_moura2021$effect.size.id <- as.factor(dat_moura2021$effect.size.id)
head(dat_moura2021)

system.time(
    phylo_eg1_tmb <- glmmTMB(yi ~ 1 + equalto(0 + effect.size.id|g, VCV) +
                         (1|study.id) + 
                         (1|species.id.phy) +
                         propto(0 + species.id.phy|g, A),
                       data = dat_moura2021,
                       REML = T)
    )
head(confint(phylo_eg1_tmb), 10)
#                                                                           2.5 %    97.5 %   Estimate
# (Intercept)                                                            0.1132610 0.6230707 0.3681658
# Std.Dev.(Intercept)|study.id                                           0.1056462 0.1813458 0.1384142
# Std.Dev.(Intercept)|species.id.phy                                     0.1932854 0.2879763 0.2359271
# Std.Dev.species.id.phyAcanthurus_leucosternon_ott388125|g.1            0.1293612 0.3959730 0.2263262
# Std.Dev.species.id.phyAcanthurus_nigricans_ott467313|g.1               0.1293612 0.3959730 0.2263262
# Std.Dev.species.id.phyAcanthurus_nigrofuscus_ott605289|g.1             0.1293612 0.3959730 0.2263262
# Std.Dev.species.id.phyAchatina_fulica_ott997087|g.1                    0.1293612 0.3959730 0.2263262
# Std.Dev.species.id.phyAegithalos_glaucogularis_vinaceus_ott5560982|g.1 0.1293612 0.3959730 0.2263262
# Std.Dev.species.id.phyAethia_pusilla_ott855484|g.1                     0.1293612 0.3959730 0.2263262
# Std.Dev.species.id.phyAgalychnis_callidryas_ott9483|g.1                0.1293612 0.3959730 0.2263262

sigma(phylo_eg1_tmb)^2
# [1] 0.01445014 # residual variance

phylo_eg1_tmb_varcor <- VarCorr(phylo_eg1_tmb)$cond
exp(phylo_eg1_tmb$fit$par)
#   betadisp      theta      theta      theta 
# 0.12020875 0.13841423 0.23592711 0.05122353 <- residual variance (sd), study.id (sd), species.id (sd), species.id.phy (variance?)

```

#### Ornstein-Uhlenbeck

```{r}
# create phylogenetic distance matrix
dat_moura2021$const <- 1
I  <- matrix(1, nrow = length(tree$tip.label), ncol = length(tree$tip.label))
D <- I-A

D[1:5, 1:5]

system.time(
    phy_ou_meta <- rma.mv(yi, vi,
   random = list(~ 1 | study.id, 
   ~ 1 | effect.size.id, 
   ~ 1 | species.id, 
   ~ species.id.phy | const),
   dist = list(D), 
   struct = "SPEXP",
   control = list(rho.init = 1),
   data = dat_moura2021,
   verbose = TRUE,
   sparse = TRUE,
   method = "REML",
   test = "t"
   )
   )

summary(phy_ou_meta)
# Multivariate Meta-Analysis Model (k = 1828; method: REML)

#    logLik   Deviance       AIC        BIC       AICc   
# -160.3783   320.7567   332.7567   365.8192   332.8028   

# Variance Components:

#             estim    sqrt  nlvls  fixed          factor 
# sigma^2.1  0.0157  0.1254    457     no        study.id 
# sigma^2.2  0.0144  0.1201   1828     no  effect.size.id 
# sigma^2.3  0.0000  0.0000    341     no      species.id 

# outer factor: const           (nlvls = 1)
# inner term:   ~species.id.phy (nlvls = 341)

#             estim    sqrt  fixed 
# tau^2      0.1029  0.3208     no 
# rho        0.0182             no 

# Test for Heterogeneity:
# Q(df = 1827) = 10743.8076, p-val < .0001

# Model Results:

# estimate      se    zval    pval   ci.lb   ci.ub      
#   0.3514  0.0355  9.9090  <.0001  0.2819  0.4209  *** 

# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


rho <- phy_ou_meta$rho  #[1] 0.01820382
alpha <- 1/rho # 54.93351
A_OU <- exp(-alpha*D)

system.time(
  phy_ou_meta1 <- rma.mv(yi, vi,
     random = list(~ 1 | study.id, 
    ~ 1 | effect.size.id, 
    ~ 1 | species.id, 
    ~ 1 | species.id.phy),
    R = list(species.id.phy = A_OU), 
    data = dat_moura2021,
    verbose = TRUE,
    sparse = TRUE,
    method = "REML",
    test = "t"
    )
)

summary(phy_ou_meta1)
# Multivariate Meta-Analysis Model (k = 1828; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -160.3783   320.7567   330.7567   358.3088   330.7896   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed          factor    R 
# sigma^2.1  0.0157  0.1254    457     no        study.id   no 
# sigma^2.2  0.0144  0.1201   1828     no  effect.size.id   no 
# sigma^2.3  0.0000  0.0001    341     no      species.id   no 
# sigma^2.4  0.1029  0.3208    341     no  species.id.phy  yes 
# 
# Test for Heterogeneity:
# Q(df = 1827) = 10743.8076, p-val < .0001
# 
# Model Results:
# 
# estimate      se    tval    df    pval   ci.lb   ci.ub      
#   0.3514  0.0355  9.9090  1827  <.0001  0.2819  0.4210  *** 

```


:::

# Spatial meta-analysis

## How to make spatial correlation matrix?

## Examples

::: panel-tabset
###1. Grau-Andrés et al. (2024)

```{r}

dat_Roger <- read.csv(here("data", "examples", "Roger_etal_2024", "Roger_etal_2024.csv"))
head(dat_Roger)

dat_Roger$const <- 1 # add constant for spatial models
dat_Roger$effect_id <- seq_len(nrow(dat_Roger))

dat_Roger <- dat_Roger %>%
  filter(!is.na(longitude))
coords2 <- cbind(dat_Roger$longitude, dat_Roger$latitude)

# project to a planar coordinate system and convert to kilometers
dat_sf2 <- st_as_sf(dat_Roger, coords = c("longitude", "latitude"), crs = 4326)
dat_sf2_proj <- st_transform(dat_sf2, crs = 3857)

coords_m <- st_coordinates(dat_sf2_proj)
dat_Roger$x_km <- coords_m[,1] / 1000
dat_Roger$y_km <- coords_m[,2] / 1000

coords_km <- cbind(dat_Roger$x_km, dat_Roger$y_km)
dist_matrix_euclid2 <- as.matrix(dist(coords_km))

# row and column names
rownames(dist_matrix_euclid2) <- dat_Roger$effect_id
colnames(dist_matrix_euclid2) <- dat_Roger$effect_id

dat_Roger <- dat_Roger %>%
  group_by(latitude, longitude) %>%
  mutate(site_id = cur_group_id()) %>%
  ungroup()
names(dat_Roger)
```

#### Exponential kernel

##### metafor
```{r}
#| eval: false
system.time(sp_eg1_exp_mf <- rma.mv(d_Hedges, var_Hedges, 
                              random = list(
                                ~ 1|effect_id,
                                # ~ 1|site_id,
                                ~ x_km + y_km |const
                              ), 
                              struct = "SPEXP", 
                              data = dat_Roger,
                              sparse = TRUE,
                              verbose = TRUE,
                              method = "REML",
                              test = "t"
                              )
            )


summary(sp_eg1_exp_mf)
# Multivariate Meta-Analysis Model (k = 2361; method: REML)
# 
#     logLik    Deviance         AIC         BIC        AICc   
# -4034.4852   8068.9704   8076.9704   8100.0360   8076.9874   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed     factor 
# sigma^2    0.7935  0.8908   2361     no  effect_id 
# 
# outer factor: const        (nlvls = 1)
# inner term:   ~x_km + y_km (nlvls = 383)
# 
#             estim    sqrt  fixed 
# tau^2      1.2330  1.1104     no 
# rho        0.1744             no 
# 
# Test for Heterogeneity:
# Q(df = 2360) = 21273.5149, p-val < .0001
# 
# Model Results:
# 
# estimate      se     tval    df    pval    ci.lb    ci.ub      
#  -0.3349  0.0644  -5.2006  2360  <.0001  -0.4612  -0.2086  *** 

```

##### brms

```{r}
#| eval: false

dat_Roger$effect_id <- as.character(dat_Roger$effect_id)
vcv <- diag(dat_Roger$var_Hedges)

rownames(vcv) <- colnames(vcv) <- dat_Roger$effect_id

fit_sp_eg1_exp_brms <- bf(d_Hedges ~ 1 +
                 # (1|site_id) + # this is site-level random effect
                 (1|gr(effect_id, cov = vcv)) + # this is m (sampling error)
                gp(x_km, y_km, 
                   cov = "exponential",
                   scale = FALSE)
                )

# generate default priors
prior <- default_prior(fit_sp_eg1_exp_brms, 
                           data = dat_Roger, 
                           data2 = list(vcv = vcv),
                           family = gaussian())
prior$prior[5] = "constant(1)" # meta-analysis assumes sampling variance is known so fixing this to 1
prior
 #                            prior     class       coef     group resp dpar nlpar lb ub tag       source
 #          student_t(3, -0.1, 2.5) Intercept                                                     default
 #                           (flat)    lscale                                       0             default
 # inv_gamma(1.494197, 2141.138175)    lscale gpx_kmy_km                            0             default
 #             student_t(3, 0, 2.5)        sd                                       0             default
 #                      constant(1)        sd            effect_id                  0             default <- check this
 #                      constant(1)        sd  Intercept effect_id                  0        (vectorized) <- check this
 #             student_t(3, 0, 2.5)      sdgp                                       0             default
 #             student_t(3, 0, 2.5)      sdgp gpx_kmy_km                            0        (vectorized)
 #             student_t(3, 0, 2.5)     sigma                                       0             default

# fitting model
sp_eg1_exp_brms <- brm(
  formula = fit_sp_eg1_exp_brms,
  data = dat_Roger,
  data2 = list(vcv=vcv),
  chains = 2,
  iter = 6000,
  warmup = 3000,
  prior = prior,
  control = list(adapt_delta=0.95, max_treedepth=15)
  )
summary(sp_eg1_exp_brms)
#  Family: gaussian 
#   Links: mu = identity 
# Formula: d_Hedges ~ 1 + (1 | gr(effect_id, cov = vcv)) + gp(x_km, y_km, cov = "exponential", scale = FALSE) 
#    Data: dat_Roger (Number of observations: 2361) 
#   Draws: 2 chains, each with iter = 6000; warmup = 3000; thin = 1;
#          total post-warmup draws = 6000
# 
# Gaussian Process Hyperparameters:
#                    Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sdgp(gpx_kmy_km)       1.14      0.07     1.00     1.29 1.00     1818     3552
# lscale(gpx_kmy_km)   128.18     26.02    85.59   187.10 1.00     2227     2567
# 
# Multilevel Hyperparameters:
# ~effect_id (Number of levels: 2361) 
#               Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sd(Intercept)     1.00      0.00     1.00     1.00   NA       NA       NA
# 
# Regression Coefficients:
#           Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept    -0.32      0.09    -0.51    -0.15 1.00     1503     2737
# 
# Further Distributional Parameters:
#       Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# sigma     0.97      0.02     0.92     1.01 1.00     3047     4199
```

##### glmmTMB
```{r}
#| eval: false

dat_Roger$effect_id <- factor(dat_Roger$effect_id)
dat_Roger$site_id <- factor(dat_Roger$site_id)

VCV <- diag(dat_Roger$var_Hedges, nrow = nrow(dat_Roger)) 
rownames(VCV)<- colnames(VCV)<- dat_Roger$effect_id
VCV[1:5, 1:5]

dat_Roger$pos <- numFactor(dat_Roger$x_km, dat_Roger$y_km)

system.time(
  sp_eg1_exp_tmb <- glmmTMB(d_Hedges ~ 1 
                   + equalto(0+effect_id|const, VCV)
                   # + (1|site_id)
                   + exp(pos+0|const),
                   data = dat_Roger, 
                   REML=TRUE)
  )
 #   user  system elapsed 
 # 45.570   1.369  47.530 

summary(sp_eg1_exp_tmb)　
# Family: gaussian  ( identity )
# Formula:          d_Hedges ~ 1 + equalto(0 + effect_id | const, VCV) + exp(pos +      0 | const)
# Data: dat_Roger
# 
#       AIC       BIC    logLik -2*log(L)  df.resid 
#    9088.5    9111.5   -4540.2    9080.5      2357 
# 
# Random effects:
# 
# Conditional model:
#  Groups   Name                                     Variance  Std.Dev. Corr                                            
#  const    effect_id1                                0.202208 0.44968                                                  
#           effect_id10                               0.166676 0.40826  0.00                                            
#           effect_id100                              0.225770 0.47515  0.00 0.00                                       
#           effect_id1000                             0.197994 0.44497  0.00 0.00 0.00                                  
#           effect_id1001                             3.676968 1.91754  0.00 0.00 0.00 0.00                             
#           effect_id1002                             0.315616 0.56180  0.00 0.00 0.00 0.00 0.00                        
#           effect_id1003                             0.305460 0.55268  0.00 0.00 0.00 0.00 0.00 0.00                   
#           effect_id1004                             0.313426 0.55984  0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           effect_id1005                             0.413540 0.64307  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           effect_id1006                             0.325726 0.57072  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...                                      ...       ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  const.1  pos(16262.6644099893,-5204.51951303405)   1.866150 1.36607                                                  
#           pos(16280.4755285163,-5116.14629455727)   1.866150 1.36607  0.00                                            
#           pos(-7965.96086752978,-5019.4733863405)   1.866150 1.36607  0.00 0.00                                       
#           pos(-7959.34359171906,-4693.06364429579)  1.866150 1.36607  0.00 0.00 0.00                                  
#           pos(-7960.45678662699,-4691.63535918108)  1.866150 1.36607  0.00 0.00 0.00 0.00                             
#           pos(-7096.61753807119,-4685.92422115392)  1.866150 1.36607  0.00 0.00 0.00 0.00 0.00                        
#           pos(-7992.73943895704,-4607.717759142)    1.866150 1.36607  0.00 0.00 0.00 0.00 0.00 0.00                   
#           pos(-7903.68384632242,-4579.4258128701)   1.866150 1.36607  0.00 0.00 0.00 0.00 0.00 0.00 0.00              
#           pos(16237.0609271069,-4578.0132445546)    1.866150 1.36607  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00         
#           pos(16304.594788628,-4572.36489569299)    1.866150 1.36607  0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00 0.00    
#           ...                                      ...       ...      ...  ...  ...  ...  ...  ...  ...  ...  ...  ...
#  Residual                                           1.864305 1.36540                                                  
# Number of obs: 2361, groups:  const, 1
# 
# Dispersion estimate for gaussian family (sigma^2): 1.86 
# 
# Conditional model:
#             Estimate Std. Error z value Pr(>|z|)    
# (Intercept) -0.40936    0.08139  -5.029 4.92e-07 ***

head(confint(sp_eg1_exp_tmb), 10) 
#                                                               2.5 %     97.5 %   Estimate
# (Intercept)                                              -0.5688798 -0.2498315 -0.4093556
# Std.Dev.pos(16262.6644099893,-5204.51951303405)|const.1   1.2417049  1.5028930  1.3660708
# Std.Dev.pos(16280.4755285163,-5116.14629455727)|const.1   1.2417049  1.5028930  1.3660708
# Std.Dev.pos(-7965.96086752978,-5019.4733863405)|const.1   1.2417049  1.5028930  1.3660708
# Std.Dev.pos(-7959.34359171906,-4693.06364429579)|const.1  1.2417049  1.5028930  1.3660708
# Std.Dev.pos(-7960.45678662699,-4691.63535918108)|const.1  1.2417049  1.5028930  1.3660708
# Std.Dev.pos(-7096.61753807119,-4685.92422115392)|const.1  1.2417049  1.5028930  1.3660708
# Std.Dev.pos(-7992.73943895704,-4607.717759142)|const.1    1.2417049  1.5028930  1.3660708
# Std.Dev.pos(-7903.68384632242,-4579.4258128701)|const.1   1.2417049  1.5028930  1.3660708
# Std.Dev.pos(16237.0609271069,-4578.0132445546)|const.1    1.2417049  1.5028930  1.3660708

sigma(sp_eg1_exp_tmb) 
# 1.365396
sp_eg1_exp_tmb_varcor <- VarCorr(sp_eg1_exp_tmb)$cond
exp(sp_eg1_exp_tmb$fit$par)
#   betadisp (efect_id)     theta (tau/tau^2)      theta (rho)
#   1.3653956               1.3660708 (1.866149)   0.2476221 

```

##### summarise results
```{r}
meta <- data.frame(model = "metafor", 
                       logLik = logLik(sp_eg1_exp_mf),
                       est = sp_eg1_exp_mf$b[[1]], 
                       se = sp_eg1_exp_mf$se[[1]], 
                       sigma2.site = sp_eg1_exp_mf$sigma2[2], ## among study variance
                       sigma2.effect = sp_eg1_exp_mf$sigma2[1], ## within study variance
                       tau2 = sp_eg1_exp_mf$tau2,
                       rho = sp_eg1_exp_mf$rho
                   )


tmb <- data.frame(model = "glmmTMB",
                   logLik = logLik(sp_eg1_exp_tmb)[1],
                   est = unlist(fixef(sp_eg1_exp_tmb))[[1]], #overall mean
                   se = as.numeric(sqrt(vcov(sp_eg1_exp_tmb)[[1]])), #overall mean SE
                   sigma2.site = (sp_eg1_exp_tmb_varcor$site_id[1]), ## among site variance estimate
                   sigma2.effect = sigma(sp_eg1_exp_tmb), ## within study variance estimate -> effect id
                   tau2 = exp(sp_eg1_exp_tmb$fit$par[3]), ##spatial variance estimate
                   rho = exp(sp_eg1_exp_tmb$fit$par[4]) ## rho
                  )

summ <- posterior_summary(
  sp_eg1_exp_brms,
  variable = c(
    "b_Intercept",
    "sigma",
    "sd_study_id__Intercept",
    "sdgp_gpx_kmy_km",
    "lscale_gpx_kmy_km"
  )
)

brms <- data.frame(
  model = "brms",
  logLik = NA_real_, 
  est = summ["b_Intercept","Estimate"],
  se  = summ["b_Intercept","Est.Error"],
  sigma2.site = (summ["sd_study_id__Intercept","Estimate"])^2,
  sigma2.effect = (summ["sigma","Estimate"])^2,
  tau2 = (summ["sdgp_gpx_kmy_km","Estimate"])^2,
  rho = summ["lscale_gpx_kmy_km","Estimate"]
)

output <- rbind(meta, tmb, brms)

knitr::kable(output)
```


#### Gaussian process

##### metafor
```{r}

```
##### brms
```{r}

```

### 2. Maire et al. (2019)
We use a dataset from the `metadat` package as our second illustrative example. In the original study, spatial correlations among sampling stations were modelled using station-level coordinates along river networks. 
However, precise latitude and longitude information for individual stations is not available in the published dataset; only the locations of study sites can be geo-referenced. 
We therefore used site-level coordinates as proxies for station-level positions. Although this approximation allows us to compare the implementation of spatial meta-analytic models across `metafor`, `brms`, and `glmmTMB`, the resulting estimates are not directly comparable to those reported by Maire et al. (2019), where station-level spatial random effects were explicitly modelled.

```{r}
dat <- dat.maire2019$dat
dmat <- dat.maire2019$dmat
dmat[1:35,1:35]

sites <- tribble(
  ~site_id, ~site_name,   ~lat,    ~lon,
  "site1",  "Belleville",  47.5,    2.65,
  "site2",  "Bugey",       45.8,    5.27,
  "site3",  "Chinon",      47.2,    0.18,
  "site4",  "Chooz",       50.1,    4.80,
  "site5",  "Civaux",      46.4,    0.65,
  "site6",  "Cruas",       44.6,    4.76,
  "site7",  "Dampierre",   47.7,    2.51,
  "site8",  "Nogent",      48.5,    3.52,
  "site9",  "St Alban",    45.3,    4.75,
  "site10", "St Laurent",  47.7,    2.53,
  "site11", "Tricastin",   44.3,    4.73
)


dat_maire2019 <- dat %>%
  left_join(sites, by = c("site" = "site_id")) |> 
  mutate(id = row_number())

dat_maire2019$const <- 1 # add constant for spatial models

# project to a planar coordinate system and convert to kilometers
dat_sf <- st_as_sf(dat_maire2019, coords = c("lon", "lat"), crs = 4326)
dat_sf_proj <- st_transform(dat_sf, crs = 3857)

coords_m <- st_coordinates(dat_sf_proj)
dat_maire2019$x_km <- coords_m[,1] / 1000
dat_maire2019$y_km <- coords_m[,2] / 1000

coords_km <- cbind(dat_maire2019$x_km, dat_maire2019$y_km)
dist_matrix_euclid <- as.matrix(dist(coords_km))

# row and column names
rownames(dist_matrix_euclid) <- dat_maire2019$id
colnames(dist_matrix_euclid) <- dat_maire2019$id

head(dat_maire2019)
```

#### Exponential kernel
##### metafor
```{r}
#| eval: false

ma_sp2_exp <- rma.mv(s1, vars1, 
             random = list(
                       ~ 1|id,
                       ~ 1|site,
                       ~ x_km + y_km | const), 
             struct = "SPEXP", 
             data = dat_maire2019,
             test = "t",
             method = "REML",
             sparse = TRUE,
             verbose = TRUE
             )

summary(ma_sp2_exp)
# Multivariate Meta-Analysis Model (k = 35; method: REML)
# 
# logLik   Deviance        AIC        BIC       AICc   
# -199.0469   398.0939   408.0939   415.7257   410.2367   
# 
# Variance Components:
#   
#   estim     sqrt  nlvls  fixed  factor 
# sigma^2.1   819.5017  28.6269     35     no      id 
# sigma^2.2  1653.5229  40.6635     11     no    site 
# 
# outer factor: const        (nlvls = 1)
# inner term:   ~x_km + y_km (nlvls = 11)
# 
# estim     sqrt  fixed 
# tau^2      8548.0233  92.4555     no 
# rho          97.3090              no 
# 
# Test for Heterogeneity:
#   Q(df = 34) = 191.4837, p-val < .0001
# 
# Model Results:
#   
#   estimate       se    zval    pval     ci.lb     ci.ub      
# 190.8255  40.9187  4.6635  <.0001  110.6263  271.0246  ***

confint(ma_sp2_exp)
# estimate  ci.lb     ci.ub 
# sigma^2.1 819.5017 0.0000 4240.3900 
# sigma.1    28.6269 0.0000   65.1183 
# 
# estimate  ci.lb      ci.ub 
# sigma^2.2 1653.5229 0.0000 21698.5229 
# sigma.2     40.6635 0.0000   147.3042 
# 
# estimate  ci.lb        ci.ub 
# tau^2 8548.0233 0.0000 >854802.3263 
# tau     92.4555 0.0000    >924.5552 
# 
# estimate   ci.lb     ci.ub 
# rho  97.3090 <9.7309 >973.0899 

```

```{r}

```

##### brms
```{r}

dat_maire2019$id <- as.character(dat_maire2019$id)
vcv <- diag(dat_maire2019$vars1)

rownames(vcv) <- colnames(vcv) <- dat_maire2019$id

fit_sp_eg2_exp_brms <- bf(s1 ~ 1 +
                 # (1|site) + # this is 
                 (1|gr(id, cov = vcv)) + # this is m (sampling error)
                gp(x_km, y_km, 
                   cov = "exponential",
                   scale = FALSE)
                )

# generate default priors
prior <- default_prior(fit_sp_eg2_exp_brms, 
                           data = dat_maire2019, 
                           data2 = list(vcv = vcv),
                           family = gaussian())
prior$prior[5] = "constant(1)" # meta-analysis assumes sampling variance is known so fixing this to 1
prior

# fitting model
sp_eg2_exp_brms <- brm(
  formula = fit_sp_eg2_exp_brms,
  data = dat_maire2019,
  data2 = list(vcv=vcv),
  chains = 2,
  threads = threading(threads_per_chain),
  iter = 15000,
  warmup = 12000,
  prior = prior,
  control = list(adapt_delta=0.95, max_treedepth=15)
)
summary(sp_eg2_exp_brms)
```

##### glmmTMB
```{r}

dat_maire2019$id <- factor(dat_maire2019$id)
VCV <- diag(dat_maire2019$vars1, nrow = nrow(dat_maire2019)) 
rownames(VCV)<- colnames(VCV)<- dat_maire2019$id
VCV[1:5, 1:5]

dat_maire2019$pos <- numFactor(dat_maire2019$x_km, dat_maire2019$y_km)

tmb_sp2_exp <- glmmTMB(
  s1 ~ 1 +
    equalto(0 + id | const, VCV) + 
    (1 | site) +
    exp(pos + 0 | const),     
  data = dat_maire2019,
  REML = TRUE
  )


head(confint(tmb_sp2_exp), 10)

sigma(tmb_sp2_exp) # for id

tmb_sp2_exp_varcor <- VarCorr(tmb_sp2_exp)$cond
exp(tmb_sp2_exp$fit$par[[4]]) # for spatial range parameter rho


```

#### Gaussian process

##### metafor
```{r}

ma_sp2_gau <- rma.mv(s1, vars1, 
                          random = list(
                            ~ 1|id,
                            ~ 1|site,
                            ~ x_km + y_km | const), 
                          struct = "SPGAU", 
                          data = dat_maire2019,
                          sparse = TRUE,
                          verbose = TRUE
)

summary(ma_sp2_gau)
# Multivariate Meta-Analysis Model (k = 35; method: REML)
# 
# logLik   Deviance        AIC        BIC       AICc   
# -199.1047   398.2094   408.2094   415.8412   410.3522   
# 
# Variance Components:
#   
#   estim     sqrt  nlvls  fixed  factor 
# sigma^2.1   805.9835  28.3898     35     no      id 
# sigma^2.2  3557.3247  59.6433     11     no    site 
# 
# outer factor: const        (nlvls = 1)
# inner term:   ~x_km + y_km (nlvls = 11)
# 
# estim     sqrt  fixed 
# tau^2      6637.4779  81.4707     no 
# rho         143.9412              no 
# 
# Test for Heterogeneity:
#   Q(df = 34) = 191.4837, p-val < .0001
# 
# Model Results:
#   
#   estimate       se    zval    pval     ci.lb     ci.ub      
# 186.2795  40.6944  4.5775  <.0001  106.5199  266.0392  *** 

confint(ma_sp2_gau)

```

##### brms
```{r}


```

##### glmmTMB
```{r}

tmb_sp2_gau <- glmmTMB(
  s1 ~ 1 +
    equalto(0 + id | const, VCV) + 
    (1 | site) +
    gau(pos + 0 | const),     
  data = dat_maire2019,
  REML = TRUE
)

head(confint(tmb_sp2_gau), 10)

tmb_sp2_gau_varcor <- VarCorr(tmb_sp2_gau)$cond
exp(tmb_sp2_gau$fit$par[[4]]) # for spatial range parameter rho
```

:::

# Software and package versions

```{r}
#| echo: false

sessionInfo()

```


# References
-   Maire, A., Thierry, E., Viechtbauer, W., & Daufresne, M. (2019). Poleward shift in large-river fish communities detected with a novel meta-analysis framework. Freshwater Biology, 64(6), 1143–1156. <https://doi.org/10.1111/fwb.13291>
-   Grau‐Andrés, R., Moreira, B., & Pausas, J. G. (2024). Global plant responses to intensified fire regimes. Global Ecology and Biogeography, 33(8), e13858. <https://doi.org/10.1111/geb.13858>
-    Viechtbauer, W., White, T., Noble, D., Senior, A., & Hamilton, W. K. (2025). metadat: Meta-analysis datasets (Version 1.5-2) [R package]. <https://github.com/wviechtb/metadat>

-    Rios Moura, R., Oliveira Gonzaga, M., Silva Pinto, N., Vasconcellos-Neto, J., & Requena, G. S. (2021). Assortative mating in space and time: Patterns and biases. Ecology Letters, 24(5), 1089–1102. https://doi.org/10.1111/ele.13690


