---
title: "Phylo spatial meta-analysis"
date: "`r Sys.Date()`"
# author: " "
format: 
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-float: true
    toc-title: "**CONTENTS**"
    output-file: "AM_worked_example.html"
    embed-resources: true
    # code-fold: true
    code-tools: 
      source: false
    number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "12"
    mainfont: "Roboto"
    monofont: "Roboto"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: united
    code-line-numbers: false
    grid:
      margin-width: 180px
crossref: 
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
  # cache: true
---

# setup

```{r}
#| label: setup
# Load required packages
rm(list = ls())

pacman::p_load(brms, MCMCglmm, metafor, metadat, tidyverse, data.table, crayon, here, ape,
               dplyr, tidyr, purrr, stringr, readr, readxl, lubridate, magrittr, janitor, 
               flextable, kableExtra, geosphere, rotl, data.table, openxlsx)

```

In this tutorial, we provide the basics of (1) phylogenetic, (2) spatial, and (3) phylogenetic-spatial meta-analysis. We show how to make phylogenetic correlation matrix and distance matrix - you can see the similarity of both and connection.

In phylogenetic meta-analysis, the Brownian Motion model (BM model) is most commonly used. This model assumes that closely related species tend to have similar traits, and it represents correlations based on the distances in a phylogenetic tree.

In contrast, spatial meta-analysis assumes that data points located closer together are more similar. It models correlation based on geographic distances, known as spatial correlation.

What is important here is the both approaches rely on distance to difine correlation. In this sencem a phylogenetic distance matrix and a spatial distance matrix share the same sttructure - they are both just distance matrices.

# phylogenetic meta-analysis

## How to make phylogenetic correlation matrix?

Before moving the each model, we show how to make phylogenetic correlation matrix. It is important to account for the shared evolutionary history of species. Most common way to do this is by incorporating a phylogenetic correlation matrix into your model. This matrix reflects how closely related species are,and can be derived in two main ways:

1.  [Using a specific phylogenetic tree](#using-a-specific-phylogenetic-tree)\
2.  [Using the `rotl` package (Open Tree of Life)](#using-the-rotl-package-open-tree-of-life)

In the following two sections, we introduce how to make the correlation matrix in two ways. The example below demonstrates how to compute a correlation matrix under the *Brownian motion process*. For the *Ornstein–Uhlenbeck process* and *Phylogenetic Gaussian Process*, please refere to their respective sections.

Note: sometimes, the species names in your data may not match those in the tree - please ensure that they are consistent. Also, it will be necessity when dropping species from the tree that are not in your dataset. In that case, you can use `TreeTools::DropTip()` / `ape::drop.tip()` (or `TreeTools::KeepTip()` / `ape::keep.tip()`)

### Using a specific phylogenetic tree {#using-a-specific-phylogenetic-tree}

If the phylogenetic relationships among the species you want to analyse are already known and you have access to a corresponding tree (e.g., from a published source — for birds, you can usually obtain one from BirdTree.org), you can easily generate a phylogenetic correlation matrix using simple codes.

```{r}
#| label: way 1 - phylometa
#| eval: false

# load necessary package
library(ape) 

# 1. read a phylogenetic tree file – supported formats include Newick (.nwk), Nexus (.nex), etc.
##   if the file contains multiple trees, it's better to start by selecting just one.
##   using ~50 trees is generally considered sufficient for phylogenetic meta-analysis.
##   (see: https://doi.org/10.1093/sysbio/syy089)
tree <- read.nexus("tree.txt")

# 2. check whether the tree is ultrametric and fully bifurcating
is.ultrametric(tree) # should return TRUE

# 3. compute the correlation matrix under BM process
A <- vcv(tree, corr = TRUE)

# 4. check the dimensions and preview a part of correlation matrix
dim(A) # returns the number of species (rows / columns)
A[1:5, 1:5] # show the top-left block of the matrix
```

### Using the `rotl` package (Open Tree of Life) {#using-the-rotl-package-open-tree-of-life}

If no tree is available (which is often the case), you can build one using the Open Tree of Life via the `rotl` package. This approach is especially useful when your meta-analysis includes species from diverse taxonomic groups - which is commonly the case.

```{r}
#| label: way 2 - phylometa
#| eval: false

# load necessary package
pacman::p_load(ape, 
               rotl,      # for accessing the OTL
               phytools,  # computing branch lengths
               stringr    # string manipulation
               ) 

# 1. get a list of unique species names from your dataaset
species_names <- unique(dat$species)

# 2. match the species names to OTL taxa
##   approximate matching is enabled to handle small name differences - e.g. 
##   you can specified the context name input - default is `All life`. you can check possible contexts `tnrs_contexts()` 
taxa <- tnrs_match_names(species,
                         context_name = "All life", 
                         do_approximate_matching = TRUE)

# 3. manually correct unmatch or incorrectly matched species
#   you can identify mismatches by checking:
# - if ott_id is NA (not matched)
# - if approximate_match is TRUE (matched with uncertainty)
# - if unique_name differs from the expected name

## e.g. replace ott_id for one mismatches - let's say it is drosophia melanogaster
tnrs_match_names("drosophila_melanogaster")
#             search_string             unique_name approximate_match     score ott_id is_synonym flags number_matches
# 1 drosophila_melanogaster Drosophila melanogaster              TRUE 0.9565217 505714      FALSE                    2
taxa$ott_id[taxa$search_string == "drosophila_melanogaster"] <- 505714

# 4. create a mapping between OTL names and original species names
##   this helps to restore originak species names later
taxon_map <- structure(taxa$search_string, names = taxa$unique_name)

# 5. retrieve the phylogenetic tree from the OTL
##   only use taxa that are present in the tree
tr <- tol_induced_subtree(ott_ids = ott_id(taxa)[is_in_tree(ott_id(taxa))])

# 6. clean up the tip labels (species names) in the tree
##   remove underscores and OTT IDs, and replace them with your original names
otl_tips <- strip_ott_ids(tr$tip.label, remove_underscores = TRUE)
tr$tip.label <- taxon_map[otl_tips]

# Add branch lengths to the tree (required to compute correlations)
tr <- compute.brlen(tr)

# 7. Compute the phylogenetic correlation matrix (Brownian motion assumption)!!
A <- vcv(tr, corr = TRUE)


# Optional ----
# standardise row/column names of the matrix
# (lowercase and use underscores instead of spaces)
rownames(A) <- tolower(gsub(" ", "_", rownames(A)))
colnames(A) <- tolower(gsub(" ", "_", colnames(A)))

# create nicely formatted tip labels for plotting
tr$tip.label <- str_replace(str_to_title(taxon_map[otl_tips]), "_", " ")

```

## example 1

```{r}
#| label: load_data - phylo
#| eval: false

Huberty_Denno_2004 <- read_csv(here("data", "Huberty_Denno_2004", "Huberty_Denno_2004_cleaned.csv"))
HubDen_tree <- read.tree(here("data", "Huberty_Denno_2004", "Huberty_Denno_2004_tree.txt"))

tip_labels <- HubDen_tree$tip.label
print(tip_labels)
unique(Huberty_Denno_2004$Herbivore)

setdiff(tip_labels, Huberty_Denno_2004$Herbivore)
# [1] "Semiothesia_sp."     "Cameraria_sp."       "Multireoides_sp."    "Phytocoris_nigrinus" "Parthenicus_sp."  "Aphalara_sp."  

Huberty_Denno_2004$Herbivore <- gsub("(?<=_sp)$", ".", Huberty_Denno_2004$Herbivore, perl = TRUE)
tips_to_keep <- intersect(Huberty_Denno_2004$Herbivore, tip_labels)
HubDen_tree_pruned <- drop.tip(HubDen_tree, setdiff(tip_labels, tips_to_keep))

setdiff(HubDen_tree_pruned$tip.label, Huberty_Denno_2004$Herbivore)

# make correlation matrix
HubDen_vcv <- vcv.phylo(HubDen_tree_pruned, corr = TRUE)

# make sure the herbivore names in the data match the tree
HubDen_2004 <- Huberty_Denno_2004 %>%
  filter(Herbivore %in% rownames(HubDen_vcv)) %>%
  mutate(Herbivore = factor(Herbivore)) 
HubDen_2004$Phylo <- HubDen_2004$Herbivore

HubDen_2004$effect_id <- seq_len(nrow(HubDen_2004))

```

Let's try to use `rotl` package

```{r}
#| echo: false
#| eval: false
# 
# species_names <- unique(Huberty_Denno_2004$Herbivore)
# taxa <- tnrs_match_names(species_names,
#                          context_name = "All life", 
#                          do_approximate_matching = TRUE)
# # Warning message:
# # Semiothesia_sp, Neodiprion_fulviceps, Cameraria_sp, Multireoides_sp, Aphalara_sp, Tachardiella_larreae, Phytocoris_sp are not matched 
# tnrs_match_names("Neodiprion fulviceps") 
# #         search_string unique_name approximate_match score ott_id is_synonym flags number_matches
# # 1 neodiprion fulviceps        <NA>                NA    NA     NA         NA    NA             NA
# # Warning message:
# # Neodiprion fulviceps are not matched 
# 
# tnrs_match_names("Tachardiella larreae")
# # search_string        unique_name approximate_match     score  ott_id is_synonym flags number_matches
# # 1 tachardiella larreae Tachardiella parva              TRUE 0.7777778 3445352      FALSE                    2
# 
# taxa$ott_id[taxa$search_string == "Tachardiella_larreae"] <- 3445352
# 
# # provided only sp 
# sp_only <- c("Semiothesia", "Cameraria", "Multireoides", 
#              "Aphalara", "Parthenicus", "Phytocoris")
# tnrs_match_names(sp_only, context_name = "All life")
# #   search_string                                 unique_name approximate_match     score  ott_id is_synonym          flags number_matches
# # 1   semiothesia                                 Demiothecia              TRUE 0.8181818 4116721      FALSE hidden, barren              2
# # 2     cameraria Cameraria (genus in kingdom Archaeplastida)             FALSE 1.0000000 1086195      FALSE                             2
# # 3  multireoides                                Multareoides              TRUE 0.9166667  250857      FALSE                             1
# # 4      aphalara                                    Aphalara             FALSE 1.0000000 3450985      FALSE                             1
# # 5   parthenicus                                 Parthenicus             FALSE 1.0000000  625583      FALSE                             1
# # 6    phytocoris                                  Phytocoris             FALSE 1.0000000  170631      FALSE                             1
# 
# taxa$ott_id[taxa$search_string == "Semiothesia_sp"] <- 4116721
# taxa$ott_id[taxa$search_string == "Cameraria_sp"] <- 1086195
# taxa$ott_id[taxa$search_string == "Multareoides_sp"] <- 250857
# taxa$ott_id[taxa$search_string == "Aphalara_sp"] <- 3450985
# taxa$ott_id[taxa$search_string == "Parthenicus_sp"] <- 625583
# taxa$ott_id[taxa$search_string == "Phytocoris_sp"] <- 170631
# 
# taxon_map <- structure(taxa$search_string, names = taxa$unique_name)
# valid_taxa <- taxa[!is.na(taxa$unique_name), ]
# taxon_map <- structure(valid_taxa$search_string, names = valid_taxa$unique_name)
# 
# tr <- tol_induced_subtree(ott_ids = ott_id(taxa)[is_in_tree(ott_id(taxa))])
# 
# otl_tips <- strip_ott_ids(tr$tip.label, remove_underscores = TRUE)
# 
# tr$tip.label <- taxon_map[otl_tips]
# 
# tr <- compute.brlen(tr)
# 
# A <- vcv(tr, corr = TRUE)
# 
# plot(tr)
```

```{r}

set.seed(42)
tr_bin <- multi2di(HubDen_tree_pruned, random = TRUE)

if (is.null(tr_bin$edge.length)) {
  tr_bin$edge.length <- rep(1, nrow(tr_bin$edge))
}

pos_min <- min(tr_bin$edge.length[tr_bin$edge.length > 0], na.rm = TRUE)
if (!is.finite(pos_min)) pos_min <- 1
tr_bin$edge.length[is.na(tr_bin$edge.length) | tr_bin$edge.length == 0] <- pos_min * 1e-6

is.ultrametric(tr_bin)

## ultrametric
tr_um <- compute.brlen(tr_bin, method = "Grafen")
any(tr_um$edge.length <= 0, na.rm = TRUE) 

## check
is.binary(tr_um)
is.ultrametric(tr_um)
plot(tr_um)

A <- vcv(tr_um, corr = TRUE)
```

### model fitting

<!-- model can actually run with ploytomy? -->

::: panel-tabset
## Brownian motion model and spatial linear model

The BM model assumes that traits evolve randomly over time. In this model, the correlation between two species depends on how much of their evolutionary history they share (i.e., the total branch length from the root to their most recent common ancestor). As the distance between species increase, the correlation decreases linearly.

This conceptually similar to a spatial model where nearby locations are more similar and ones are less so - also with a linear decay in correlation.

### Brownian motion model

```{r}
#| label: phylo_brownian - metafor
#| eval: false

HD_BM <- rma.mv(d, var,
                 random = list(~ 1 | Reference, 
                               ~ 1 | effect_id, 
                               ~ 1 | Herbivore, 
                               ~ 1 | Phylo),
                 R = list(Phylo = A), 
                 data = HubDen_2004)

summary(HD_BM)

# Multivariate Meta-Analysis Model (k = 85; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -161.2370   322.4741   332.4741   344.6282   333.2433   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed     factor    R 
# sigma^2.1  0.8191  0.9051     45     no  Reference   no 
# sigma^2.2  0.2244  0.4737     85     no  effect_id   no 
# sigma^2.3  1.1564  1.0754     54     no  Herbivore   no 
# sigma^2.4  2.0725  1.4396     54     no      Phylo  yes 
# 
# Test for Heterogeneity:
# Q(df = 84) = 706.6128, p-val < .0001
# 
# Model Results:
# 
# estimate      se     zval    pval    ci.lb   ci.ub    
#  -0.5027  0.8614  -0.5835  0.5595  -2.1910  1.1857    

# phylogenetic heritability

phylo <- HD_BM$sigma2[4]/ (HD_BM$sigma2[3]+HD_BM$sigma2[4]) 
phylo*100

# change the variable name (sub-guild) to sub_guild using data.table package

setDT(HubDen_2004)
setnames(HubDen_2004, old = "Sub-guild", new = "Sub_guild")

HD_BM_2 <- rma.mv(yi = d,
                    V = var,      
                    mods  = ~ -1 + Sub_guild,
                    random = list(
                      ~ 1 | Reference,
                      ~ 1 | effect_id,
                      ~ 1 | Herbivore,
                      ~ 1 | Phylo
                      ),
                    R = list(Phylo = A),
                    data  = HubDen_2004
                    )

summary(HD_BM_2)
# Multivariate Meta-Analysis Model (k = 85; method: REML)
# 
#    logLik   Deviance        AIC        BIC       AICc   
# -144.5514   289.1028   311.1028   337.0266   315.1028   
# 
# Variance Components:
# 
#             estim    sqrt  nlvls  fixed     factor    R 
# sigma^2.1  1.0431  1.0213     45     no  Reference   no 
# sigma^2.2  0.1866  0.4320     85     no  effect_id   no 
# sigma^2.3  0.8676  0.9315     54     no  Herbivore   no 
# sigma^2.4  1.8756  1.3695     54     no      Phylo  yes 
# 
# Test for Residual Heterogeneity:
# QE(df = 78) = 618.0104, p-val < .0001
# 
# Test of Moderators (coefficients 1:7):
# QM(df = 7) = 13.8791, p-val = 0.0534
# 
# Model Results:
# 
#                       estimate      se     zval    pval    ci.lb   ci.ub    
# Sub_guildBorer          0.8938  1.1735   0.7617  0.4462  -1.4061  3.1938    
# Sub_guildFree-living   -0.3839  0.9614  -0.3993  0.6897  -2.2682  1.5005    
# Sub_guildGall Former   -1.8786  1.1051  -1.6999  0.0892  -4.0446  0.2874  . 
# Sub_guildMesophyll     -1.1245  1.2825  -0.8768  0.3806  -3.6381  1.3892    
# Sub_guildMiner          1.4165  1.1358   1.2471  0.2124  -0.8097  3.6427    
# Sub_guildPhloem        -1.3502  1.0540  -1.2810  0.2002  -3.4159  0.7156    
# Sub_guildXylem         -1.0210  1.8943  -0.5390  0.5899  -4.7337  2.6917    
# 
# ---
```

In `brms`, variance is specified using the `se()` function, and we need to specify the phylogenetic covariance matrix using the `data2` argument. The `gr()` function is used to specify the phylogenetic random effect. Do not forget to specify `|a|`- it means that correlation is allowed between the phylogenetic random effects.


```{r}
#| label: dat1 - phylo_brownian - brms
#| eval: false

formula_dat1_BM <- bf(d | se(sqrt(var)) ~ 1 + 
                        (1 | Reference) + 
                        (1 | effect_id) + 
                        (1 | Herbivore) +  # non-phylogenetic random effect
                        (1 |a| gr(Phylo, cov = A))
                      )

prior_dat1_BM <- get_prior(
  formula = formula_dat1_BM,
  data = HubDen_2004,
  data2 = list(A = A),
  family = gaussian()
)

dat1_BM_brms <- brm(
  formula = formula_dat1_BM,
  family = gaussian(),
  data = HubDen_2004,
  data2 = list(A = A),
  prior = prior_dat1_BM,
  iter = 6000, 
  warmup = 4000,  
  thin = 1,
  chains = 2, 
  control = list(adapt_delta = 0.95)
)
summary(dat1_BM_brms)

```

Both the Bayesian model (`brms`) and the frequentist model (`metafor`) consistently showed strong phylogenetic signal, with the phylogenetic variance component being the largest among all random effects. The estimates of random effect magnitudes were broadly similar between the two models.

### Spatial linear model

```{r}

# add a constant column for specifying a shared correlation structure
HubDen_2004$Const <- 1

#  make distance matrix
## by subtracting the correlation matrix from the identity matrix (i.e., distance = 1 - correlation)
I <- matrix(1, nrow = 54, ncol = 54)

## convert the phylogenetic correlation matrix into a distance matrix
D_HD <- I - A 

HD_spatial_lin <- rma.mv(d, var,
                     random = list(~ 1 | Reference,  # random effect for study
                                   ~ 1 | effect_id, # random effect for each effect size
                                   ~ 1 | Herbivore, # random effect for species
                                   ~ Phylo | Const), # random effect for phylogeny
                     dist=list(D_HD),  # phylogenetic distance matrix
                     struct = "SPEXP", # use spatial exponential correlation structure
                     control=list(rho.init = 1), 
                     data = HubDen_2004)

summary(HD_spatial_lin)
summary(HD_BM)
```

## Ornstein-Uhlenbeck model

The OU model extends the BM model by assuming that traits are pulled toward an optimal value over time (this reflects stabilising selection). In the OU model, correlation between species decays exponentially with distance:

$
\text{cor}(t) = exp(-\alpha_t)
$

Here, $t$ is the phylogenetic distance and $\alpha$ controls how strongly traits are pulled toward the optimum. Larger values of $\alpha$ lead to faster decay in correlation. Interestingly, this has the same form as the exponential spatial correlation function:

$
\text{cor}(d) = exp(-d/\rho)
$

This means that the OU model and exponential spatial model are mathematically equivalent.
If we set $\alpha = 1/\rho$, the two models produce the same correlation values for the same distance.
In this way, by recognising the structural similarity between phylogenetic and spatial distance matrices, we can use the spatial modelling framework to implement the OU model in  phylogenetic meta-analysis.

This allows us to incorporate more realistic assumptions, such as stabilising selection, than the BM model. And while estimating $\alpha$ directly in OU model is often challenging, estimating $\rho$ in a spatial model is typically more straightforward (you can see the below). Thus, hacking spatial meta-analysis framework is a practical way to estimating $\alpha$ values in the meta-analysis with OU model.

:::

::: panel-tabset
### Spatial exponential model

```{r}
#| label: phylo_spatialEXP - metafor
#| eval: false

D_HD <- I - HubDen_vcv 

HD_spatial <- rma.mv(d, var,
                     random = list(~ 1 | Reference,  # random effect for study
                                   ~ 1 | effect_id, # random effect for each effect size
                                   ~ 1 | Herbivore, # random effect for species
                                   ~ Phylo | Const), # random effect for phylogeny
                     dist=list(D_HD),  # phylogenetic distance matrix
                     struct = "SPEXP", # use spatial exponential correlation structure
                     control=list(rho.init = 1), 
                     data = HubDen_2004)

summary(HD_spatial)

```
The estimated variance for the phylogenetic effect is 2.55.

$\rho$, the decay parameter of the exponential correlation function, is analogous to $1/\alpha$ in an OU model. A lower rho implies faster decay (stronger stabilising selection); a higher $\rho$ means slower decay (weaker pull).

```{r}


```

### Ornstein-Uhlenbeck model

Using the rho parameter from the spatial exponential model, we can fit the Ornstein-Uhlenbeck model.

```{r}
#| label: phylo_ornstein - metafor
#| eval: false

rho <- HD_spatial$rho
alpha <- 1/rho
HD_OU <- exp(-alpha * D_HD)

HD_OU <- rma.mv(d, var,
                random = list(~ 1 | Reference, 
                              ~ 1 | effect_id, 
                              ~ 1 | Herbivore, 
                              ~ 1 | Phylo),
                  R = list(Phylo = HD_OU), 
                  data = HubDen_2004
                )

summary(HD_OU)

```

You can see that the estimate of the Ornstein-Uhlenbeck model is the same to the spatial exponential model, but the AIC is lower for the Ornstein-Uhlenbeck model.

## Gaussian phylogenetic process model

Actually, we can fit the Gaussian phylogenetic process model using spatial Gaussian process model.

```{r}
#| label: phylo_gaussian - metafor
#| eval: false

HD_GP <- rma.mv(d, var,
                random = list(~ 1 | Reference, 
                              ~ 1 | effect_id, 
                              ~ 1 | Herbivore, 
                          ~ Phylo | Const),
                     dist=list(D_HD), 
                     struct = "SPGAU",
                     control=list(rho.init=1), 
                     data = HubDen_2004)

summary(HD_GP)

```
:::

## example 2

```{r}
#| label: load_data - example 2
#| eval: false

dat2 <- read_csv(here("data", "Hoeksema_etal_2010", "HoeksemaEtal2010EcoLett.csv"))
tree_dat2 <- read.tree(here("data", "Hoeksema_etal_2010", "HoeksemaEtal2010EcoLett_tree.txt"))

# make species column merge using "PlantGenus" and "PlantSpecificEpithet" columns (i.e., "PlantGenus_PlantSpecificEpithet") - but all characters should be lower case
dat2$species <- paste0(tolower(dat2$PlantGenus), "_", tolower(dat2$PlantSpecificEpithet))

# check whether the the tip labels and species name of data2 match

setdiff(tree_dat2$tip.label, dat2$species)
setdiff(dat2$species, tree_dat2$tip.label)

# it looks like the unmatched cases in tip labels and species names can be fixed?
tree_dat2$tip.label[grep("fragaria_", tree_dat2$tip.label)]
tree_dat2$tip.label[grep("_cacao", tree_dat2$tip.label)]
dat2$species[grep("fragaria_", dat2$species)] <- "fragaria_sp"
dat2$species[grep("theobrama_cacao", dat2$species)] <- "theobroma_cacao" # typo - raw data has "theobrama cacao" but the tree has "theobroma_cacao"

# remove the species "NA_NA" rows
dat2 <- dat2 %>% 
  filter(!species %in% c("NA_NA"))

```

Then, we need to create the correlation matrix from the phylogenetic tree for phylogenetic meta-analysis.

```{r}
#| eval: false

# make variance covariance matrix

dat2_vcv <- vcv.phylo(tree_dat2, corr = TRUE) # make sure the `corr = TRUE`
dat2_vcv[1:5, 1:5] # check the first few rows and columns

```

### model fitting

::: panel-tabset
## Brownian motion model

Let's fit the Brownian motion model using the `rma.mv` function from the `metafor`and `brms` packages. First, we run the metafor model, then we run the brms model.

```{r}
#| label: dat2 - phylo_brownian - metafor
#| eval: false

# check variable names
names(dat2)

# change the variable names, and also add effect_id/phylo column
dat2 <- dat2 %>%
  rename(
    lnRR = `Effect_size (natural log response ratio of mean inoculated plant mass over mean non-inoculated plant mass)`,
    var = `Estimated_variance (inverse of summed replicates of inoculated and non-inoculated plants)`,
    study_id = Pub_1st_author
  )
dat2$effect_id <- seq_len(nrow(dat2))
dat2$phylo <- dat2$species

dat2_BM_metafor <- rma.mv(lnRR, var,
                 random = list(~ 1 | study_id, 
                               ~ 1 | effect_id, 
                               ~ 1 | species, 
                               ~ 1 | phylo),
                 R = list(phylo = dat2_vcv), 
                 data = dat2)

summary(dat2_BM_metafor)

```

In `brms`, variance is specified using the `se()` function, and we need to specify the phylogenetic covariance matrix using the `data2` argument. The `gr()` function is used to specify the phylogenetic random effect. Do not forget to specify `|a|`- it means that correlation is allowed between the phylogenetic random effects.

```{r}
#| label: dat2 - phylo_brownian - brms
#| eval: false

formula_dat2_BM <- bf(lnRR | se(sqrt(var)) ~ 1 + 
                        (1 | study_id) + 
                        (1 | effect_id) + 
                        (1 | species) +  # non-phylogenetic random effect
                        (1 |a| gr(phylo, cov = dat2_vcv))
                      )

prior_dat2_BM <- get_prior(
  formula = formula_dat2_BM,
  data = dat2,
  data2 = list(dat2_vcv = dat2_vcv),
  family = gaussian()
)

dat2_BM_brms <- brm(
  formula = formula_dat2_BM,
  family = gaussian(),
  data = dat2,
  data2 = list(dat2_vcv = dat2_vcv),
  prior = prior_dat2_BM,
  iter = 6000, 
  warmup = 4000,  
  thin = 1,
  chains = 2, 
  control = list(adapt_delta = 0.95)
  
)
summary(dat2_BM_brms)

```

## Ornstein-Uhlenbeck model

It is a bit tricky when we fit the OU model - we need to calculate the $\alpha$ parameter from the rho parameter, and then create the correlation matrix using the `exp(-alpha * D)` formula, where `D` is the distance matrix. To do that, we first fit the spatial exponential model, then we use the $\rho$ parameter to the alpha parameter, and finally we create the correlation matrix. <!-- I am trying how to do that using the `brms` package, but it is not working yet... -->

```{r}
#| label: dat2 - phylo_ornstein - metafor
#| eval: false

dat2$const <- 1

# make identity matrix to calculate the distance matrix
I <- matrix(1, nrow = nrow(dat2_vcv), ncol = ncol(dat2_vcv))

# then, make distance matrix
D_dat2 <- I - dat2_vcv 

dat2_spatial <- rma.mv(lnRR, var,
                     random = list(~ 1 | study_id, 
                                   ~ 1 | effect_id, 
                                   ~ 1 | species, 
                                   ~ phylo | const),
                     dist=list(D_dat2), 
                     struct = "SPEXP",
                     control=list(rho.init=1), 
                     data = dat2)

summary(dat2_spatial)


```

Then, we can extract the $\rho$ parameter and calculate the $\alpha$ parameter, and then create the correlation matrix using the `exp(-alpha * D)` formula.

```{r}
#| label: dat2 - phylo_ornstein - metafor - alpha
#| eval: false

rho_dat2 <- dat2_spatial$rho
alpha_dat2 <- 1/rho_dat2
dat2_OU <- exp(-alpha_dat2 * D_dat2)

dat2_OU <- rma.mv(lnRR, var,
                random = list(~ 1 | study_id, 
                              ~ 1 | effect_id, 
                              ~ 1 | species,  
                              ~ 1 | phylo),
                  R = list(phylo = dat2_OU), 
                  data = dat2
                )

summary(dat2_OU)

```

## Gaussian phylogenetic process model

```{r}
#| label: dat2 - phylo_gaussian - metafor
#| eval: false

dat2_GP <- rma.mv(lnRR, var,
                     random = list(~ 1 | study_id, 
                                   ~ 1 | effect_id, 
                                   ~ 1 | species, 
                                   ~ phylo | const),
                     dist=list(D_dat2), 
                     struct = "SPGAU",
                     control=list(rho.init=1), 
                     data = dat2)

summary(dat2_GP)


```

Let's compare the all results. You can find the estimates are the same, but the AIC is lower for the OU model..

```{r}
#| label: dat2 - phylo_ornstein - metafor - compar
#| eval: false

summary(dat2_spatial)
summary(dat2_OU)
summary(dat2_BM_metafor)
summary(dat2_GP)


## make summary table ----

## make list
model_list <- list(
  "BM_metafor" = dat2_BM_metafor,
  "OU_metafor" = dat2_OU,
  "Spatial_EXP_metafor" = dat2_spatial,
  "GP_metafor" = dat2_GP
)

m <- dat2_BM_metafor
m <- model_list$SPEXP
m$tau2
m$rho
lapply(model_list, function(x) is.na(x$rho))
lapply(model_list, function(x) is.na(x$tau2))

## function for tidying up the model results ----

tidy_table <- function(m) {

  library("tidyr")
  
  # extract the sigma2 vector (random effect variances),
  # transpose it to a row, and convert to a data frame
  sigmas <- m$sigma2 |> t() |> as.data.frame()
  
  # assign names to the sigma2 columns using m$s.names
  # these correspond to random effect levels - "study", "species", "phylo", etc.
  names(sigmas) <- m$s.names
  
  # use broom::tidy() to extract model summary statistics
  # then add:
  # - lower and upper 95% confidence intervals (from m$ci.lb and m$ci.ub)
  # - rho (e.g., spatial/phylogenetic correlation parameter)
  # - tau2 (variance in spatial/GP models - spatial model patameter)
  # - AIC (model fit)
  
  m.tidy <- tidy(m) %>% 
    mutate(
      lwr_ci = m$ci.lb,
      upper_ci = m$ci.ub,
      rho = m$rho,
      tau2 = m$tau2,
      AIC = AIC(m)
    ) %>%
    # bind the sigma2 values as columns to the tidy output
    bind_cols(sigmas)
  
  # convert the result to a data.table for easier manipulation if needed
  setDT(m.tidy)
  
  # return the complete tidy table
  return(m.tidy)
}

# run the tidy_table function on each model in the model_list
tidy.models <- lapply(model_list, tidy_table)

# combine the tidy models into a single data.table
tidy.models.dat <- rbindlist(tidy.models,
                             fill = TRUE,
                             idcol = "model_id")
tidy.models.dat
```
:::

# Spatial meta-analysis

## Example 1

```{r}
#| label: load_data - spatial
#| eval: false

dat_coetzee <- read.xlsx(here("data", "examples", "Coetzee_2014.xlsx"), sheet = 2)
dat_coetzee$const <- 1 # add constant for spatial models

# make distance matrix
coords <- cbind(dat_coetzee$long, dat_coetzee$lat)
dist_matrix <- distm(coords, fun = distHaversine)
dist_matrix[1:5,1:5]

rownames(dist_matrix) <- dat_coetzee$datapt_id
colnames(dist_matrix) <- dat_coetzee$datapt_id

```

::: panel-tabset
## spatial exponential model

```{r}
#| label: spatial_exp - metafor
#| eval: false

EXP_mod1 <- rma.mv(yi, vi, 
                  random = list(
                        ~ 1|datapt_id,
                        ~ 1|study_id, 
                        ~ lat + long | const
                      ), 
                    struct="SPEXP", 
                    data = dat_coetzee,
                    # dist=list(dist_matrix),
                    # control=list(rho.init = 10)
                  )

summary(EXP_mod1)
```

### brms

```{r}

fit_1 <- bf(yi | se(sqrt(vi)) ~ 1 + 
                        (1 | datapt_id) + 
                        (1 | study_id) + 
                        gp(long, lat, cov = "exponential"))

prior <- get_prior(formula = fit_1,
                   data = dat_coetzee1,
  family = gaussian()
  )

m_exp <- brm(formula = fit_1,
             data = dat_coetzee1,
             family = gaussian(),
             prior = prior,
             iter = 8000,
             warmup = 6000,
             chain = 2, 
             thin = 1
             )

summary(m_exp)
```

## spatial Gaussian process model

```{r}
#| label: spatial_gaussian - metafor
#| eval: false

EXP_mod2 <- rma.mv(yi, vi, 
                      random = list(
                        ~ 1|datapt_id,
                        ~ 1|study_id, 
                        ~ lat + long | const
                      ), 
                      struct="SPGAU", 
                      data = dat_coetzee)

summary(EXP_mod2)
```

```{r}
fit_2 <- bf(yi | se(sqrt(vi)) ~ 1 + 
                        (1 | datapt_id) + 
                        (1 | study_id) + 
                        gp(long, lat, cov = "exp_quad"))

prior2 <- get_prior(formula = fit_2,
                   data = dat_coetzee,
  family = gaussian()
  )

m_gp <- brm(formula = fit_2,
             data = dat_coetzee,
             family = gaussian(),
             prior = prior2,
             iter = 15000,
             warmup = 13000,
             chain = 2, 
             thin = 1
             )
summary(m_gp)
```

## spatial linear model

```{r}
#| label: spatial_linear - metafor
#| eval: false

EXP_mod3 <- rma.mv(yi, vi, 
                      random = list(
                        ~ 1|datapt_id,
                        ~ 1|study_id, 
                        ~ datapt_id | const
                      ), 
                      struct="SPLIN", 
                      data = dat_coetzee, 
                      dist=list(dist_matrix),
                      control=list(rho.init = 10))

summary(EXP_mod3)
```
:::

## Example 2

# phylo-spatial meta-analysis

Current version, the model estimated no the phylogenetic effect and spatial effect - I need to find ideal dataset (I can detect such effects) to test this model.

```{r}
#| label: phylo_spatial - metafor
#| eval: false

library(rotl)
library(readxl)
library(here)

data <- read_excel(here("data", "examples", "Coetzee_2014.xlsx"), sheet = 2)
data$spp_name[data$spp_name == "NA"] <- NA

df$species <- tolower(gsub(" ", "_", df$spp_name))
species_list <- unique(df$species)

taxa <- tnrs_match_names(species_list)

taxon_map <- structure(taxa$search_string, names = taxa$unique_name)

tr <- tol_induced_subtree(ott_id(taxa)[is_in_tree(ott_id(taxa))])

otl_tips <- strip_ott_ids(tr$tip.label, remove_underscores = TRUE)
tr$tip.label <- taxon_map[otl_tips]

tr$node.label <- NULL
tr <- compute.brlen(tr)
cor <- vcv(tr, corr = TRUE)

tr_corrected <- tr
tr_corrected$tip.label <- str_replace(str_to_title(taxon_map[otl_tips]), "_", " ")

df <- df[df$species %in% rownames(cor), ]
cor <- cor[df$species, df$species]

df$const <- 1 

coords <- as.matrix(df[, c("long", "lat")])
D_geo <- distm(coords, fun = distHaversine)
rownames(D_geo) <- colnames(D_geo) <- as.character(df$datapt_id)
names(df)

spatial_phylo_mod1 <- rma.mv(
  yi = yi,
  V = vi,
  random = list(
    ~ 1|datapt_id,
    ~ 1|study_id, 
    ~ datapt_id | const,
    ~ 1 | species
  ),
  R = list(species = cor),
  struct="SPEXP", 
  data = df, 
  dist = list(D_geo),
  # control = list(rho.init = 10)
)

summary(spatial_phylo_mod1)

spatial_phylo_mod0 <- rma.mv(
  yi = yi,
  V = vi,
  random = list(
    ~ 1|datapt_id,
    ~ 1|study_id
  ),
  data = df
)
summary(spatial_phylo_mod0)

# spatial
spatial_mod <- rma.mv(
  yi = yi,
  V = vi,
  random = list(
    ~ 1|datapt_id,
    ~ 1|study_id, 
    ~ datapt_id|const
  ),
  struct="SPEXP", 
  data = df, 
  dist = list(D_geo)
)

summary(spatial_mod)

# phylo
phylo_mod <- rma.mv(
  yi = yi,
  V = vi,
  random = list(
    ~ 1|datapt_id,
    ~ 1|study_id, 
    ~ 1|species
  ),
  R = list(species = cor),
  data = df
)

summary(phylo_mod)

AIC(spatial_phylo_mod0, spatial_phylo_mod1, spatial_mod, phylo_mod)

```
