---
title: "Model 2 (count data)"
date: "`r Sys.Date()`"
# author: " "
format: 
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-float: true
    toc-title: "**CONTENTS**"
    output-file: "AM_worked_example.html"
    embed-resources: true
    # code-fold: true
    code-tools: 
      source: false
    number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "12"
    mainfont: "Roboto"
    monofont: "Roboto"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: united
    code-line-numbers: false
    grid:
      margin-width: 180px
crossref: 
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
  # cache: true
---

# setup
```{r}
#| label: setup
# Load required packages

pacman::p_load(brms, MCMCglmm, metafor, metadat, tidyverse, dplyr, tidyr, purrr, stringr, readr, readxl, lubridate, magrittr, janitor, flextable, kableExtra, geosphere, rtol, here)


```

# phylogenetic meta-analysis

```{r}
#| label: load_data - phylo

Huberty_Denno_2004 <- read_csv(here("data", "Huberty_Denno_2004", "Huberty_Denno_2004_cleaned.csv"))
HubDen_tree <- read.tree(here("data", "Huberty_Denno_2004", "Huberty_Denno_2004_tree.txt"))

tip_labels <- HubDen_tree$tip.label
print(tip_labels)
unique(Huberty_Denno_2004$Herbivore)

setdiff(tip_labels, Huberty_Denno_2004$Herbivore)
# [1] "Semiothesia_sp."     "Cameraria_sp."       "Multireoides_sp."    "Phytocoris_nigrinus" "Parthenicus_sp."    
# [6] "Aphalara_sp."  

Huberty_Denno_2004$Herbivore <- gsub("(?<=_sp)$", ".", Huberty_Denno_2004$Herbivore, perl = TRUE)
tips_to_keep <- intersect(Huberty_Denno_2004$Herbivore, tip_labels)
HubDen_tree_pruned <- drop.tip(HubDen_tree, setdiff(tip_labels, tips_to_keep))

setdiff(HubDen_tree_pruned$tip.label, Huberty_Denno_2004$Herbivore)

# make variance covariance matrix
HubDen_vcv <- vcv.phylo(HubDen_tree_pruned, corr = TRUE)

# make sure the herbivore names in the data match the tree
HubDen_2004 <- Huberty_Denno_2004 %>%
  filter(Herbivore %in% rownames(HubDen_vcv)) %>%
  mutate(Herbivore = factor(Herbivore)) 
HubDen_2004$Phylo <- HubDen_2004$Herbivore

HubDen_2004$effect_id <- seq_len(nrow(HubDen_2004))

```

## model fitting

::: panel-tabset
## Brownian motion model

```{r}
#| label: phylo_brownian - metafor

HD_BM <- rma.mv(d, var,
                 random = list(~ 1 | Reference, 
                               ~ 1 | effect_id, 
                               ~ 1 | Herbivore, 
                               ~ 1 | Phylo),
                 R = list(Phylo = HubDen_vcv), 
                 data = HubDen_2004)

summary(HD_BM)
```

## Ornstein-Uhlenbeck model
To fit an Ornstein-Uhlenbeck model, we need to calculate the  

### spatial exponential model
```{r}
#| label: phylo_spatialEXP - metafor
HubDen_2004$Const <- 1
I <- matrix(1, nrow = 54, ncol = 54)

D_HD <- I - HubDen_vcv # make distance matrix

HD_spatial <- rma.mv(d, var,
                     random = list(~ 1 | Reference, 
                                   ~ 1 | effect_id, 
                                   ~ 1 | Herbivore, 
                                   ~ Phylo | Const),
                     dist=list(D_HD), 
                     struct = "SPEXP",
                     control=list(rho.init=1), 
                     data = HubDen_2004)

summary(HD_spatial)

```

### Ornstein-Uhlenbeck model

Using the rho parameter from the spatial exponential model, we can fit an Ornstein-Uhlenbeck model.
```{r}
#| label: phylo_ornstein - metafor

rho <- HD_spatial$rho
alpha <- 1/rho
HD_OU <- exp(-alpha * D_HD)

HD_OU <- rma.mv(d, var,
                random = list(~ 1 | Reference, 
                              ~ 1 | effect_id, 
                              ~ 1 | Herbivore, 
                              ~ 1 | Phylo),
                  R = list(Phylo = HD_OU), 
                  data = HubDen_2004
                )

summary(HD_OU)

```
You can see that the estimate of the Ornstein-Uhlenbeck model is the same to the spatial exponential model, but the AIC is lower for the Ornstein-Uhlenbeck model.


## Gaussian phylogenetic process model
Actually, we can fit the Gaussian phylogenetic process model using spatial Gaussian process model.

```{r}
#| label: phylo_gaussian - metafor

HD_GP <- rma.mv(d, var,
                random = list(~ 1 | Reference, 
                              ~ 1 | effect_id, 
                              ~ 1 | Herbivore, 
                          ~ Phylo | Const),
                     dist=list(D_HD), 
                     struct = "SPGAU",
                     control=list(rho.init=1), 
                     data = HubDen_2004)

summary(HD_GP)

```

:::

# spatial meta-analysis

```{r}
#| label: load_data - spatial

dat_coetzee <- read.xlsx(here("data", "examples", "Coetzee_2014.xlsx"), sheet = 2)
dat_coetzee$const <- 1 # add constant for spatial models

# make distance matrix
coords <- cbind(dat_coetzee$long, dat_coetzee$lat)
dist_matrix <- distm(coords, fun = distHaversine)
dist_matrix[1:5,1:5]

rownames(dist_matrix) <- dat_coetzee$datapt_id
colnames(dist_matrix) <- dat_coetzee$datapt_id

```

::: panel-tabset

## spatial exponential model
```{r}
#| label: spatial_exp - metafor

EXP_mod1 <- rma.mv(yi, vi, 
                  random = list(
                        ~ 1|datapt_id,
                        ~ 1|study_id, 
                        ~ datapt_id | const
                      ), 
                    struct="SPEXP", 
                    data = dat_coetzee, 
                    dist=list(dist_matrix),
                    control=list(rho.init = 10))

summary(EXP_mod1)
```

## spatial Gaussian process model

```{r}
#| label: spatial_gaussian - metafor

EXP_mod2 <- rma.mv(yi, vi, 
                      random = list(
                        ~ 1|datapt_id,
                        ~ 1|study_id, 
                        ~ datapt_id | const
                      ), 
                      struct="SPGAU", 
                      data = dat_coetzee, 
                      dist=list(dist_matrix),
                      control=list(rho.init=10))

summary(EXP_mod2)
```

## spatial linear model

```{r}
#| label: spatial_linear - metafor

EXP_mod3 <- rma.mv(yi, vi, 
                      random = list(
                        ~ 1|datapt_id,
                        ~ 1|study_id, 
                        ~ datapt_id | const
                      ), 
                      struct="SPLIN", 
                      data = dat_coetzee, 
                      dist=list(dist_matrix),
                      control=list(rho.init = 10))

summary(EXP_mod3)
```
:::

# phylo-spatial meta-analysis
Current version, the model estimated no the phylogenetic effect and spatial effect - I need to find ideal dataset (I can detect such effects) to test this model.
```{r}
#| label: phylo_spatial - metafor

library(rotl)
library(readxl)
library(here)

data <- read_excel(here("data", "examples", "Coetzee_2014.xlsx"), sheet = 2)
data$spp_name[data$spp_name == "NA"] <- NA

df$species <- tolower(gsub(" ", "_", df$spp_name))
species_list <- unique(df$species)

taxa <- tnrs_match_names(species_list)

taxon_map <- structure(taxa$search_string, names = taxa$unique_name)

tr <- tol_induced_subtree(ott_id(taxa)[is_in_tree(ott_id(taxa))])

otl_tips <- strip_ott_ids(tr$tip.label, remove_underscores = TRUE)
tr$tip.label <- taxon_map[otl_tips]

tr$node.label <- NULL
tr <- compute.brlen(tr)
cor <- vcv(tr, corr = TRUE)

tr_corrected <- tr
tr_corrected$tip.label <- str_replace(str_to_title(taxon_map[otl_tips]), "_", " ")

df <- df[df$species %in% rownames(cor), ]
cor <- cor[df$species, df$species]

df$const <- 1 

coords <- as.matrix(df[, c("long", "lat")])
D_geo <- distm(coords, fun = distHaversine)
rownames(D_geo) <- colnames(D_geo) <- as.character(df$datapt_id)
names(df)

spatial_phylo_mod1 <- rma.mv(
  yi = yi,
  V = vi,
  random = list(
    ~ 1|datapt_id,
    ~ 1|study_id, 
    ~ datapt_id | const,
    ~ 1 | species
  ),
  R = list(species = cor),
  struct="SPEXP", 
  data = df, 
  dist = list(D_geo),
  # control = list(rho.init = 10)
)

summary(spatial_phylo_mod1)

spatial_phylo_mod0 <- rma.mv(
  yi = yi,
  V = vi,
  random = list(
    ~ 1|datapt_id,
    ~ 1|study_id
  ),
  data = df
)
summary(spatial_phylo_mod0)

# spatial
spatial_mod <- rma.mv(
  yi = yi,
  V = vi,
  random = list(
    ~ 1|datapt_id,
    ~ 1|study_id, 
    ~ datapt_id|const
  ),
  struct="SPEXP", 
  data = df, 
  dist = list(D_geo)
)

summary(spatial_mod)

# phylo
phylo_mod <- rma.mv(
  yi = yi,
  V = vi,
  random = list(
    ~ 1|datapt_id,
    ~ 1|study_id, 
    ~ 1|species
  ),
  R = list(species = cor),
  data = df
)

summary(phylo_mod)

AIC(spatial_phylo_mod0, spatial_phylo_mod1, spatial_mod, phylo_mod)

```