---
title: "Model 2 (count data)"
date: "`r Sys.Date()`"
# author: " "
format: 
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-float: true
    toc-title: "**CONTENTS**"
    output-file: "AM_worked_example.html"
    embed-resources: true
    # code-fold: true
    code-tools: 
      source: false
    number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "12"
    mainfont: "Roboto"
    monofont: "Roboto"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: united
    code-line-numbers: false
    grid:
      margin-width: 180px
crossref: 
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: "—"     # (default is ":")
  fig-prefix: Fig.     # (default is "Figure")
  tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
  # cache: true
---

# setup
```{r}
#| label: setup
# Load required packages
rm(list = ls())

pacman::p_load(brms, MCMCglmm, metafor, metadat, tidyverse, data.table, crayon, 
               dplyr, tidyr, purrr, stringr, readr, readxl, lubridate, magrittr, janitor, 
               flextable, kableExtra, geosphere, rotl, here)

```

# phylogenetic meta-analysis
## example 1
```{r}
#| label: load_data - phylo

Huberty_Denno_2004 <- read_csv(here("data", "Huberty_Denno_2004", "Huberty_Denno_2004_cleaned.csv"))
HubDen_tree <- read.tree(here("data", "Huberty_Denno_2004", "Huberty_Denno_2004_tree.txt"))

tip_labels <- HubDen_tree$tip.label
print(tip_labels)
unique(Huberty_Denno_2004$Herbivore)

setdiff(tip_labels, Huberty_Denno_2004$Herbivore)
# [1] "Semiothesia_sp."     "Cameraria_sp."       "Multireoides_sp."    "Phytocoris_nigrinus" "Parthenicus_sp."  "Aphalara_sp."  

Huberty_Denno_2004$Herbivore <- gsub("(?<=_sp)$", ".", Huberty_Denno_2004$Herbivore, perl = TRUE)
tips_to_keep <- intersect(Huberty_Denno_2004$Herbivore, tip_labels)
HubDen_tree_pruned <- drop.tip(HubDen_tree, setdiff(tip_labels, tips_to_keep))

setdiff(HubDen_tree_pruned$tip.label, Huberty_Denno_2004$Herbivore)

# make variance covariance matrix
HubDen_vcv <- vcv.phylo(HubDen_tree_pruned, corr = TRUE)

# make sure the herbivore names in the data match the tree
HubDen_2004 <- Huberty_Denno_2004 %>%
  filter(Herbivore %in% rownames(HubDen_vcv)) %>%
  mutate(Herbivore = factor(Herbivore)) 
HubDen_2004$Phylo <- HubDen_2004$Herbivore

HubDen_2004$effect_id <- seq_len(nrow(HubDen_2004))

```

### model fitting

::: panel-tabset
## Brownian motion model

```{r}
#| label: phylo_brownian - metafor

HD_BM <- rma.mv(d, var,
                 random = list(~ 1 | Reference, 
                               ~ 1 | effect_id, 
                               ~ 1 | Herbivore, 
                               ~ 1 | Phylo),
                 R = list(Phylo = HubDen_vcv), 
                 data = HubDen_2004)

summary(HD_BM)
```

## Ornstein-Uhlenbeck model

### spatial exponential model
```{r}
#| label: phylo_spatialEXP - metafor

HubDen_2004$Const <- 1
I <- matrix(1, nrow = 54, ncol = 54)

D_HD <- I - HubDen_vcv # make distance matrix

HD_spatial <- rma.mv(d, var,
                     random = list(~ 1 | Reference, 
                                   ~ 1 | effect_id, 
                                   ~ 1 | Herbivore, 
                                   ~ Phylo | Const),
                     dist=list(D_HD), 
                     struct = "SPEXP",
                     control=list(rho.init=1), 
                     data = HubDen_2004)

summary(HD_spatial)

```

### Ornstein-Uhlenbeck model

Using the rho parameter from the spatial exponential model, we can fit an Ornstein-Uhlenbeck model.
```{r}
#| label: phylo_ornstein - metafor

rho <- HD_spatial$rho
alpha <- 1/rho
HD_OU <- exp(-alpha * D_HD)

HD_OU <- rma.mv(d, var,
                random = list(~ 1 | Reference, 
                              ~ 1 | effect_id, 
                              ~ 1 | Herbivore, 
                              ~ 1 | Phylo),
                  R = list(Phylo = HD_OU), 
                  data = HubDen_2004
                )

summary(HD_OU)

```
You can see that the estimate of the Ornstein-Uhlenbeck model is the same to the spatial exponential model, but the AIC is lower for the Ornstein-Uhlenbeck model.


## Gaussian phylogenetic process model
Actually, we can fit the Gaussian phylogenetic process model using spatial Gaussian process model.

```{r}
#| label: phylo_gaussian - metafor

HD_GP <- rma.mv(d, var,
                random = list(~ 1 | Reference, 
                              ~ 1 | effect_id, 
                              ~ 1 | Herbivore, 
                          ~ Phylo | Const),
                     dist=list(D_HD), 
                     struct = "SPGAU",
                     control=list(rho.init=1), 
                     data = HubDen_2004)

summary(HD_GP)

```

:::

## example 2
```{r}
#| label: load_data - example 2

dat2 <- read_csv(here("data", "Hoeksema_etal_2010", "HoeksemaEtal2010EcoLett.csv"))
tree_dat2 <- read.tree(here("data", "Hoeksema_etal_2010", "HoeksemaEtal2010EcoLett_tree.txt"))

# make species column merge using "PlantGenus" and "PlantSpecificEpithet" columns (i.e., "PlantGenus_PlantSpecificEpithet") - but all characters should be lower case
dat2$species <- paste0(tolower(dat2$PlantGenus), "_", tolower(dat2$PlantSpecificEpithet))

# check whether the the tip labels and species name of data2 match

setdiff(tree_dat2$tip.label, dat2$species)
setdiff(dat2$species, tree_dat2$tip.label)

# it looks like the unmatched cases in tip labels and species names can be fixed?
tree_dat2$tip.label[grep("fragaria_", tree_dat2$tip.label)]
tree_dat2$tip.label[grep("_cacao", tree_dat2$tip.label)]
dat2$species[grep("fragaria_", dat2$species)] <- "fragaria_sp"
dat2$species[grep("theobrama_cacao", dat2$species)] <- "theobroma_cacao" # typo - raw data has "theobrama cacao" but the tree has "theobroma_cacao"

# remove the species "NA_NA" rows
dat2 <- dat2 %>% 
  filter(!species %in% c("NA_NA"))

```

Then, we need to create the correlation matrix from the phylogenetic tree for phylogenetic meta-analysis. 
```{r}
# make variance covariance matrix

dat2_vcv <- vcv.phylo(tree_dat2, corr = TRUE) # make sure the `corr = TRUE`
dat2_vcv[1:5, 1:5] # check the first few rows and columns

```

### model fitting

::: panel-tabset

## Brownian motion model

Let's fit the Brownian motion model using the `rma.mv` function from the `metafor`and `brms` packages.
First, we run the metafor model, then we run the brms model.
```{r}
#| label: dat2 - phylo_brownian - metafor

# check variable names
names(dat2)

# change the variable names, and also add effect_id/phylo column
dat2 <- dat2 %>%
  rename(
    lnRR = `Effect_size (natural log response ratio of mean inoculated plant mass over mean non-inoculated plant mass)`,
    var = `Estimated_variance (inverse of summed replicates of inoculated and non-inoculated plants)`,
    study_id = Pub_1st_author
  )
dat2$effect_id <- seq_len(nrow(dat2))
dat2$phylo <- dat2$species

dat2_BM_metafor <- rma.mv(lnRR, var,
                 random = list(~ 1 | study_id, 
                               ~ 1 | effect_id, 
                               ~ 1 | species, 
                               ~ 1 | phylo),
                 R = list(phylo = dat2_vcv), 
                 data = dat2)

summary(dat2_BM_metafor)

```

In `brms`, variance is specified using the `se()` function, and we need to specify the phylogenetic covariance matrix using the `data2` argument. The `gr()` function is used to specify the phylogenetic random effect. Do not forget to specify `|a|`- it means that correlation is allowed between the phylogenetic random effects.

```{r}
#| label: dat2 - phylo_brownian - brms

formula_dat2_BM <- bf(lnRR | se(sqrt(var)) ~ 1 + 
                        (1 | study_id) + 
                        (1 | effect_id) + 
                        (1 | species) +  # non-phylogenetic random effect
                        (1 |a| gr(phylo, cov = dat2_vcv))
                      )

prior_dat2_BM <- get_prior(
  formula = formula_dat2_BM,
  data = dat2,
  data2 = list(dat2_vcv = dat2_vcv),
  family = gaussian()
)

dat2_BM_brms <- brm(
  formula = formula_dat2_BM,
  family = gaussian(),
  data = dat2,
  data2 = list(dat2_vcv = dat2_vcv),
  prior = prior_dat2_BM,
  iter = 6000, 
  warmup = 4000,  
  thin = 1,
  chains = 2, 
  control = list(adapt_delta = 0.95)
  
)
summary(dat2_BM_brms)

```

## Ornstein-Uhlenbeck model
It is a bit tricky when we fit the OU model - we need to calculate the $\alpha$ parameter from the rho parameter, and then create the correlation matrix using the `exp(-alpha * D)` formula, where `D` is the distance matrix. To do that, we first fit the spatial exponential model, then we use the $\rho$ parameter to the alpha parameter, and finally we create the correlation matrix.
<!-- I am trying how to do that using the `brms` package, but it is not working yet... -->

```{r}
#| label: dat2 - phylo_ornstein - metafor

dat2$const <- 1

# make identity matrix to calculate the distance matrix
I <- matrix(1, nrow = nrow(dat2_vcv), ncol = ncol(dat2_vcv))

# then, make distance matrix
D_dat2 <- I - dat2_vcv 

dat2_spatial <- rma.mv(lnRR, var,
                     random = list(~ 1 | study_id, 
                                   ~ 1 | effect_id, 
                                   ~ 1 | species, 
                                   ~ phylo | const),
                     dist=list(D_dat2), 
                     struct = "SPEXP",
                     control=list(rho.init=1), 
                     data = dat2)

summary(dat2_spatial)


```

Then, we can extract the $\rho$ parameter and calculate the $\alpha$ parameter, and then create the correlation matrix using the `exp(-alpha * D)` formula.

```{r}
#| label: dat2 - phylo_ornstein - metafor - alpha


rho_dat2 <- dat2_spatial$rho
alpha_dat2 <- 1/rho_dat2
dat2_OU <- exp(-alpha_dat2 * D_dat2)

dat2_OU <- rma.mv(lnRR, var,
                random = list(~ 1 | study_id, 
                              ~ 1 | effect_id, 
                              ~ 1 | species,  
                              ~ 1 | phylo),
                  R = list(phylo = dat2_OU), 
                  data = dat2
                )

summary(dat2_OU)

```

Let's compare the results from spatial exponential model and OU model. You can find the estimates are the same, but the AIC is lower for the OU model..

```{r}
#| label: dat2 - phylo_ornstein - metafor - compar

summary(dat2_spatial)
summary(dat2_OU)
summary(dat2_BM_metafor)
AIC(dat2_spatial, dat2_OU, dat2_BM_metafor)

```

## Gaussian phylogenetic process model
```{r}
#| label: dat2 - phylo_gaussian - metafor

dat2_GP <- rma.mv(lnRR, var,
                     random = list(~ 1 | study_id, 
                                   ~ 1 | effect_id, 
                                   ~ 1 | species, 
                                   ~ phylo | const),
                     dist=list(D_dat2), 
                     struct = "SPGAU",
                     control=list(rho.init=1), 
                     data = dat2)

summary(dat2_GP)


```

:::

# spatial meta-analysis

```{r}
#| label: load_data - spatial

dat_coetzee <- read.xlsx(here("data", "examples", "Coetzee_2014.xlsx"), sheet = 2)
dat_coetzee$const <- 1 # add constant for spatial models

# make distance matrix
coords <- cbind(dat_coetzee$long, dat_coetzee$lat)
dist_matrix <- distm(coords, fun = distHaversine)
dist_matrix[1:5,1:5]

rownames(dist_matrix) <- dat_coetzee$datapt_id
colnames(dist_matrix) <- dat_coetzee$datapt_id

```

::: panel-tabset

## spatial exponential model
```{r}
#| label: spatial_exp - metafor

EXP_mod1 <- rma.mv(yi, vi, 
                  random = list(
                        ~ 1|datapt_id,
                        ~ 1|study_id, 
                        ~ datapt_id | const
                      ), 
                    struct="SPEXP", 
                    data = dat_coetzee, 
                    dist=list(dist_matrix),
                    control=list(rho.init = 10))

summary(EXP_mod1)
```

## spatial Gaussian process model

```{r}
#| label: spatial_gaussian - metafor

EXP_mod2 <- rma.mv(yi, vi, 
                      random = list(
                        ~ 1|datapt_id,
                        ~ 1|study_id, 
                        ~ datapt_id | const
                      ), 
                      struct="SPGAU", 
                      data = dat_coetzee, 
                      dist=list(dist_matrix),
                      control=list(rho.init=10))

summary(EXP_mod2)
```

## spatial linear model

```{r}
#| label: spatial_linear - metafor

EXP_mod3 <- rma.mv(yi, vi, 
                      random = list(
                        ~ 1|datapt_id,
                        ~ 1|study_id, 
                        ~ datapt_id | const
                      ), 
                      struct="SPLIN", 
                      data = dat_coetzee, 
                      dist=list(dist_matrix),
                      control=list(rho.init = 10))

summary(EXP_mod3)
```
:::

# phylo-spatial meta-analysis
Current version, the model estimated no the phylogenetic effect and spatial effect - I need to find ideal dataset (I can detect such effects) to test this model.
```{r}
#| label: phylo_spatial - metafor

library(rotl)
library(readxl)
library(here)

data <- read_excel(here("data", "examples", "Coetzee_2014.xlsx"), sheet = 2)
data$spp_name[data$spp_name == "NA"] <- NA

df$species <- tolower(gsub(" ", "_", df$spp_name))
species_list <- unique(df$species)

taxa <- tnrs_match_names(species_list)

taxon_map <- structure(taxa$search_string, names = taxa$unique_name)

tr <- tol_induced_subtree(ott_id(taxa)[is_in_tree(ott_id(taxa))])

otl_tips <- strip_ott_ids(tr$tip.label, remove_underscores = TRUE)
tr$tip.label <- taxon_map[otl_tips]

tr$node.label <- NULL
tr <- compute.brlen(tr)
cor <- vcv(tr, corr = TRUE)

tr_corrected <- tr
tr_corrected$tip.label <- str_replace(str_to_title(taxon_map[otl_tips]), "_", " ")

df <- df[df$species %in% rownames(cor), ]
cor <- cor[df$species, df$species]

df$const <- 1 

coords <- as.matrix(df[, c("long", "lat")])
D_geo <- distm(coords, fun = distHaversine)
rownames(D_geo) <- colnames(D_geo) <- as.character(df$datapt_id)
names(df)

spatial_phylo_mod1 <- rma.mv(
  yi = yi,
  V = vi,
  random = list(
    ~ 1|datapt_id,
    ~ 1|study_id, 
    ~ datapt_id | const,
    ~ 1 | species
  ),
  R = list(species = cor),
  struct="SPEXP", 
  data = df, 
  dist = list(D_geo),
  # control = list(rho.init = 10)
)

summary(spatial_phylo_mod1)

spatial_phylo_mod0 <- rma.mv(
  yi = yi,
  V = vi,
  random = list(
    ~ 1|datapt_id,
    ~ 1|study_id
  ),
  data = df
)
summary(spatial_phylo_mod0)

# spatial
spatial_mod <- rma.mv(
  yi = yi,
  V = vi,
  random = list(
    ~ 1|datapt_id,
    ~ 1|study_id, 
    ~ datapt_id|const
  ),
  struct="SPEXP", 
  data = df, 
  dist = list(D_geo)
)

summary(spatial_mod)

# phylo
phylo_mod <- rma.mv(
  yi = yi,
  V = vi,
  random = list(
    ~ 1|datapt_id,
    ~ 1|study_id, 
    ~ 1|species
  ),
  R = list(species = cor),
  data = df
)

summary(phylo_mod)

AIC(spatial_phylo_mod0, spatial_phylo_mod1, spatial_mod, phylo_mod)

```